### 2018-05-18 DraftReport.tex
first upload

\documentclass{article}
\usepackage{latexsym, amssymb, amsmath, amsfonts, bbm, graphicx,float}
\usepackage{caption}
\DeclareCaptionLabelFormat{cont}{#1~#2\alph{ContinuedFloat}}
\captionsetup[ContinuedFloat]{labelformat=cont}
\usepackage{subcaption}
\usepackage[margin=2.5cm]{geometry}
%\usepackage{natbib}[authoryear]
%\setcitestyle{authoryear,open={((},close={))}}
\author{Jennifer Israelsson}

\begin{document}
	\title{Interim report}
	\maketitle

	\section{Introduction}
	In many places of the world, countries depend on agriculture for food and export. Predicting and modelling rainfall is therefore an important skill, that many researchers have attempted all over the world. Many reserachers have proposed different probability distributions, with the most common ones being: gamma, lognormal, exponential and Weibull (Shariffah et al, 2007) (Suhaila et al, 2011) (Stern et al, 1984). Instead of trying to fit the data to just one distribution(Shariffah et al, 2007) worked with a mixture of two distributions of the same type but with different parameters and a probability parameter to give the two distributions different weight. To determine the most suitable distribution function, either various goodness-of-fit tests are used or Akaike information criterion(Suhaila et al, 2011). Commonly used goodness-of-fit tests are: Kolmogorov-Smirnov, Cramer-von-Mises and Anderson-Darling. The problem with using several goodness-of-fit tests is that they might indicate different distribution fits as the most suitible one, since their scores focuses on different measures. The authors of (Shariffah et al, 2007) did not weight the importance of the tests differently, but choose to rank them after how many tests the specific distribution did better in. In both (Shariffah et al, 2007) and (Suhaila et al, 2011 )the authors concluded that the daily rainfall distriution was strongly connected to the surrounding topography. The authors of (Shariffah et al, 2007) also slit the data into moonsoon season and transition periods to see if the differents seasons followed different distributions, and concluded that it was a big difference between the stations depending on if they were on the east coast or the west coast. \\
	To model the occurence of rain, two common ideas is to use a binomial distribution with p equal to the number of wet days divided by the total number of days, or a Markov chain. With a Markov chain, one calculates the probability of having a pattern of a desired length, e.g what is the probability of having a rainy day if it has rained for the past two days. The problem with a Markov chain is that it assumes stationarity, i.e that the probabilities are the same for any time of the year. This is not true in most parts of the world due to seasonality. The authors in (Stern et al, 1984) instead lets the probabilites depend on a link function $h$ and some other function $g$, which depends on the rain occurence for the past days. This function g takes the form of a Fourier series and the number of harmonics is determined by using multiple regression techniques and stops including terms when no improvement is acheived by including another term. By using this method, they can use the full data set and let the probabilities vary over the year. To improve the fit even further, the authors split the occurence of rain into days of trace rain, defined as rain less then 2.5 mm, and rainy days.\\
	A different statistical approach is used in (Sanso et al, 1999), where the authors, instead of trying to fit a model to the data points, treat the data points as realisations of a multivariate normal distribution, with each station having a different month dependent mean. The data they are working with is measured in mm and does not classify any rain as "traces". This method tries to predict rainfall in both space and time, by letting the observation be a realisation of a model depending on both. They base their model on a truncated normal model
	\begin{equation*}
	z_{it} =
	\begin{cases}
	w^{\beta_t}_{it},& \text{if} \ w_{it}> 0 \\
	0, & \text{if} \ w_{it}  \leq 0 
	\end{cases}
	\end{equation*}
	where $z_{it}$ is the observed rainfall at station $i$ = $1,...,k$ and month $t$, $\beta_t > 0$ and $w$ is a normal random variable. By letting $w_t = (w_{1t}, ...,w_{kt})$, $w_t \sim N_k(\mu_t, \sigma^2_t\Sigma)$, i.e $w_t$ is a multivariate normal variable with mean vector $\mu_t$ and covariance matrix $\Sigma$. The authors let the mean depend on some spatial parameters and the entries of the covariance matrix is defined as $\Sigma_{ij} = \text{exp}(-\lambda d_{ij})$, where $\lambda >$ 0 and $d_{ij}$ is the distance between station i and j after they are projected onto a horizontal plane. By using a Bayesian method, they incorporate the uncertainty of the parameters into the posterior distribution. \\

	Ghana is a country in west Africas by the coast and shares boarder with Burkina Faso, C\^{o}te d'Ivoire and Togo. It has five distinct geographical area: low plains in the south, the Volta Basin in the centre with the artificial lake 'Lake Volta', the Akwapim-Togo ranges to the east of the Volta Basin with many heights and folded strata, the Ashanti Uplands to the west and high plains in the north (FAO, 2005). The temperature is peaking around February-March and at its lowest around August. Ghana has three distinct rainfall behvaiours. The northern part experiences an unimodal season with the rainy season between April and September, wheras the rest of the country has a bimodal season, first one in April to July and the second in September to November. The difference is that some parts of the country has two modes of the same amplitude wheras others have peaks of different amplitudes. But they all have in common a slowly increasing peak but a rapid decrease in October(Nkrumah et al, 2014). The different rain patterns depend on a few wind and pressure phenomenons. It is strongly affected by the  position of the \textbf{Inter-tropcial convergence zone}, which goes between the nothern and southern tropics every year. As ITCZ moves from the nothern position to the southern and back, the opposing prevailing winds gives rise to the West African monsoon which shows as the two rainy seasons. The rainy season in the north corresponds to when the ITCZ are at its most nothern position.The prevailing winds north of the ITCZ is called the Harmmatan and brings hot and dusty air from the Sahara desert between Dacember and March, which gives rise to the very dry season. The prevailing wind south of ITCZ is southwesterly and instead brings humid air from the Atlantic ocean.\\

	In this project I will start to look at if it is resonable to fit the entire data set in each month, after dividing the data into groups depending on rainfall pattern, to a single distribution or if you can get a more accurate distribution fit if only using values up to a certain value. To investigate this I will first use Quntile-Quantile ,QQ, plots to find probable threshold values and then use goodness-of-fit tests to see if there has been an improvment in the fit.

	\section{Probability}
	The most commonly used distributions for daily rainfall amounts are gamma, lognormal, exponential and Weibull, eventhought the last one is an extreme value distribution. In this report, my first attempt has been to fit the data to a gamma distribution by Maximum likelihood estimation, MLE, and tried to fit the data to the other distibutions mentioned if I did not get a good fit based on QQ plots. MLE is a method of choosing parameters that maximizes the likelihood function, by differentiating the log liklihood fuction with respect to the parameter to be estimated, and then equate it to zero, we obtain the parameter. This method gives the smallest variance comapred to other parameter estimation methods such as method of moments and generalized probability weighted moments (Shariffah et al, 2007). \\
	\textbf{Gamma distribution: } The gamma distribution is a two-parameter continuous distribution family, characterized by a shape parameter, $\alpha$, and a scale parameter, $\theta$, both positive. For a random variable X the probability distribution function,PDF, mean and variance is given by,
	\begin{equation*}
	f(x ; \alpha, \theta) = \frac{x^{\alpha-1} e^{-\frac{x}{\theta}}}{\theta^\alpha \Gamma(\alpha)}, \ x > 0
	\end{equation*}
	\begin{center}
		$E(X) = \alpha\theta$, $Var[X] = \alpha\theta^2$
	\end{center}
	\textbf{Weibull: }The Weibull distribution is a two parameter distribution depending on a shape parameter $k$ and a scale parameter $\lambda$, both positive. For a random variable X the PDF, mean and variance is given by,
	\begin{equation*}
	f(x ; k, \lambda) = \frac{k}{\lambda}\left(\frac{x}{\lambda}\right)^{k-1}e^{(-x/ \lambda)^k}, \ x> 0
	\end{equation*}
	\begin{equation*}
	E(X) = \lambda\Gamma\left(1+\frac{1}{k}\right), \ Var[X] = \lambda^2\left[\Gamma\left(1 + \frac{2}{k}\right)- \left(\Gamma\left(1 + \frac{1}{k}\right)\right)^2\right]
	\end{equation*}
	\subsection{Statistical test}
	To compare different fitted distributions and determine the most suitible ones, one can use both statistical and graphical methods. Graphical methods, such as QQ plots, offers a fast way of comparing your data to a fitted model or looking at general behaviour of the data, wheras statistical tools quantifies how similar two distributions are two each other by calculating various measures of distance such as maximum distance between distribution functions. The issue with applying goodness-of-fit tests to large data sets is that your p-values can be very small, hence suggest you to reject your null hypothesis, when in reality the difference is very small. It is therfore difficult to use these test to determine if the fitted model is suitable, but can be used to compare the fit of two different models if they have a similar size on their data sets.

	\subsubsection{Statistical tools}
	\textbf{Wilcoxon signed rank test:} Is a non-parametric test that either tests the null hypothesis that a sample has a certain median or the null hypothesis that two paired samples comes from the same distribution. In the paired case, you take the difference between each pair, rank the difference from smallest to largest absolut value and then add up all the ranks of the positive differences. If this value is greater then the table value for our number of pairs, we reject the null hypothesis and conclude that the data comes from different distributions. 
	 This test can be used to test if the median in months has significantly changed over a time period, given that they have he same pattern. If we compare two sets, one sample with the peak in April and one sample with the peak in September, they will clearly behave differently but the test will give the same weight to the positive and the negative rankings, so we will not reject our null hypothesis. \linebreak
	\textbf{Kolmogorov-Smirnov test:} This test is used to test if our emperical cumulative distribution function and our assumed theoretical one are close enough. The test statistic is given by,
	\begin{equation*}
	D_n = sup_x |F_n(x) - F(x)|
	\end{equation*}
	where $F_n$ is our empirical distrbution function, EDF, and $F$ is the cumulative distribution function, CDF, that we are testing against. We reject our null hypothesis if $D_n$ is larger then our table value.\\
	\textbf{Anderson-Darling test:} Both this test and the Cram\'{e}r-von Mises test belongs to the class of quadratic EDF statistics, where this test puts bigger weight on the tail behaviour. The test statistics, A, of our ordered data ${x_1, x_2, ..., x_n}$, is defined as,
	\begin{equation*}
	A^2 = -n - S
	\end{equation*}
	where
	\begin{equation*}
	S = \sum_{i=1}^n\frac{2i -1}{n}[\text{ln}(F(x_i)) + \text{ln}(1 - F(x_{n+1-i}) )]
	\end{equation*}
	where $F$ is the CDF that we are testing our data against. \\
	\textbf{Cram\'{e}r-von Mises:} This test is similar to the Anderson-Darling test. We once again test our increasingly ordered data ${x_1, x_2,...,x_n}$ with a CDF $F$ with the test statistic,
	\begin{equation*}
	T = \frac{1}{12n} + \sum_{i=1}^{n}\left[\frac{2i-1}{2n} - F(x_i)\right]^2
	\end{equation*}
	and reject our null hypothesis if the T value is larger then the table value.\\
	\textbf{Akaike information criterion:} Is an estimator of how good a model is compared to other models, fitted to the same data set. The number does not tell if the model is a good fit or not, only if it is a better fit then some other model. The best model fit is the one with the lowest value. 
	\subsubsection{Graphical methods}
	\textbf{Quantile-Quantile plots, QQ}: A QQ  plot compares the quantiles of our data, with the qunatiles of our assumed distribution. If these two distributions agree, our data will fall on an approxiamatly straight line. If our data falls below the straight line, our assumed distribution is more heavy tailed then our data, meaning that our data has fewer higher values then expected. If our data insteads falls above the straight line, our assumed distribution is too light tailed. \\
	\textbf{Autocorrelation plot:} Autocorrelation describes how correlated the time series data is to itself, i.e how dependent is the data a few days forward on the day today. By plotting correlation as a function on number of lag days, we can see roughly how long a wet spell is in each month. For autocorrelation to work, there can be no missing values, so one needs to remove the months from the data with missing values. 
	\subsection{Extreme value theory}
	\subsubsection{Maximum analysis}
	In extreme value theory, we only look at the larger values in our data instead of all the data, to get a better understanding of how the tail behaves, and gives us a more accurate way to extrapolate outside our data. We divide our data into suitible blocks, e.g years or months, and denote $M_n$ = max$(X_1,...,X_n)$, i.e the maximum value of our n sample points in each block. This is often called a block maxima. We are interested in finding the distribution of $M_n$ as n increases. We call a non-degenerated rv X \textit{max-stable} if
	\begin{equation*}
	c_nX + d_n = M_n
	\end{equation*}
	for all n $\geq$ 2 and appropriate $c_n>0$, $d_n \in \mathbb{R}$. Under the assumption that all our $X_i$'s are independent and identically distributed, we also have that
	\begin{equation*}
	\mathbb{P}(M_n \leq x) = \mathbb{P}(X_1 \leq x,...,X_n \leq x) = F^n(x)
	\end{equation*}
	with F being the distribution function of the rv X. This leads us to the important \textit{Fisher-Tippet theorem}, which states that; if there exists norming constants $c_n$ and $d_n$ as above and some non-degenerated distribution function H such that
	\begin{equation}
	c_n^{-1}(M_n - d_n) \xrightarrow{d} H
	\end{equation}
	then H belongs to one of the three extreme value distributions; Fr\'{e}chet, Gumbel or Weibull. These can be summarised in one distribution, called \textit{Generalised extreme value} distribution, GEV. 
	\begin{equation*}
	H_\xi(x) = 
	\begin{cases}
	\text{exp}(-(1+\xi x)^{-1/\xi}), & \text{if} \ \xi \neq 0 \\
	\text{exp}(-(\text{exp}(-x))), & \text{if} \ \xi = 0 
	\end{cases}
	\end{equation*}
	where $1+\xi x > 0$. With $\xi = 0$, we get Gumbel, $\xi > 0$ Fr\'{e}chet and $\xi < 0$ Weibull. Closely related to GEV distributions is \textit{maximum domain of attraction}. We say that a rv X belongs to the \textit{maximum domain of attraction} of \textit{H}, X $\in MDA(H)$, if there exists norming constants $c_n$ $>$ 0, $d_n \in \mathbb{R}$ such that (1) holds.

	\subsubsection{Upper order analysis}
	Instead of looking at maxima, and thereby throwing away a lot of data, we can look at all values above a threshold, u. We can look at both the distribution and the mean of the data above this threshold
	\begin{equation*}
	F_u(x) = \mathbb{P}(X - u \leq x | X > u) = \frac{F(x + u) - F(u)}{1 - F(u)}
	\end{equation*}
	\begin{equation*}
	e(u) = \mathbb{E}(X - u | X > u)
	\end{equation*}
	$e(u)$ is called \textit{the mean excess function of X}. Just like the block maxima converges to a GEV distribution, the data over a high threshold also converges to a distribution, called the \textit{General pareto distribution}, GPD
	\begin{equation*}
	G_{\xi, \nu, \sigma}(x) =
	\begin{cases}
	1-(1+\xi \frac{x-\nu}{\sigma})^{-1/\xi}, & \text{if} \ 1+\xi \frac{x-\nu}{\sigma} > 0, \xi \neq 0 \\
	1- e^{-\frac{x-\nu}{\sigma}}, & \text{if} \  \xi = 0
	\end{cases}
	\end{equation*}
	where
	\begin{equation*}
	\begin{cases}
	x \geq 0, & \text{if} \ \xi \geq 0 \\
	0 \leq x \leq -\sigma/ \xi, & \text{if} \ \xi < 0
	\end{cases}
	\end{equation*}
	$\nu \in \mathbb{R}$ is called a location parameter and $\sigma>0$ is a scale parameter. The two distributions are related, such that the distributions for which the block maxima converges to a GEV with parameter $\xi$, their excess distribution converges to the GPD with same shape parameter $\xi$.

	\section{Data}
	The data that will be used in this report is 30 years of daily precipitation data from 21 stations around Ghana. A overview of the data is displayed in table \ref{Tstations}. Because of many missing values, the stations BEK, ABE and AKA has been excluded from further analysis, and station NAV and WA has been excluded because their data sets are close to identical to TLE.
	\subsection{Overall data structure}

	\begin{table}[H]
		\centering
		\begin{tabular}{| c | c | c | c | c| c| c | c | c |}
			\hline
			Station & Long. & Lat. & Pos. Obs. & Missing obs. & Days (Dec, Jan, Feb) & Annual mean & Rainy day mean & Mode \\
			\hline
			AXM & -2.23 & 4.86 & 4282 & 58 & 214, 104, 154 & 1874 & 13.13 &  Semi-bi \\
			ODA & -0.98 & 5.93 & 4180 & 0 & 140, 60, 151 & 1407 & 10.10 & Bi \\
			BEK & -2.33 & 6.2 & 3851 & 366 & 80, 6, 141 & 1394 & 10.50 & \\
			KDA & -0.25 & 6.08 & 3756 & 30 & 112 , 81, 167 & 1293 & 10.33 & Bi\\
			ABE & -0.73 & 6.65 & 3415 & 700& 65, 46, 112 &1277 & 10.47 &  \\
			KSI & -1.6 & 6.71 & 3586 & 0 & 80 , 51, 136 & 1347 & 11.27 & Bi \\
			HO & 0.46 & 6.6 & 3409 & 0& 86, 52, 140 & 1276 & 11.23& Bi \\
			TDI & -1.76 & 4.88 & 3312 & 28 & 113, 37, 92 & 1079 & 9.77 & Semi-bi\\
			SUN & -2.33 & 7.33 & 3187 & 61 & 52, 23, 106 & 1191 & 11.22 & Bi\\
			WEN & -2.1 & 7.75 & 3188 & 2& 48, 24, 90 & 1249 & 11.75& Bi \\
			KRA & -0.03 & 7.81 & 2991 & 0 & 30, 20, 35 & 1366 & 13.70 & Uni\\
			BOL & -2.48 & 9.03 & 2688 & 2 &  21, 7, 29 & 1101 & 12.29 & Uni\\
			SAL & -1.06 & 5.2 & 2679 & 123 & 52, 23, 106 & 931 & 10.43 & Semi-bi\\
			NAV & -0.01 & 9.45 & 2623 & 62& 6, 6, 22 & 1024 & 11.71 & \\
			WA & -2.5 & 10.1 & 2609 & 62 & 6, 6, 22 & 1018 & 11.70 &  \\
			TLE & -0.85 & 8.5 & 2599 & 31 & 7, 6, 23 & 1017 & 11.74 & Uni\\
			AKA & 0.8 & 6.11 & 2236 & 580 & 74, 23, 56 & 848 & 10.99 & \\
			NAV1 & -1.1 & 10.9 & 2184 & 32 & 4, 2, 11 & 988 & 13.57 & Uni\\
			ACC & -0.16 & 5.6 & 2130 & 6 & 54, 29, 56 & 747 & 10.52 & Semi-bi \\
			ADA & 0.63 & 5.78 & 2037 & 0 & 37, 21, 58 & 790 & 11.63 & Semi-bi\\
			TEM & 0 & 5.61 & 1840 & 31 & 42, 23, 49 & 659 & 10.75 & Semi-bi \\
			\hline
		\end{tabular}
		\caption{All stations}
		\label{Tstations}
	\end{table}

	By plotting the monthly average of each station, one can easily see a clear uni-, bi- and semibimodal behaviour for each station. I have therefore grouped the stations into these three categories and plot these 3 groups monthly averages instead, to get even more data to work with (fig \ref{MonAvegroups}). Plotting the stations on a map, colour coded by their rainmode (fig \ref{mapRainmode}), one can clearly see the pattern of only one rainy season in the north and two in the south, with a stronger rainy season in April for the coast station and two even rainy seasons for the central stations. This confirms the pattern that others have observed and explained with the shift in the ITCZ.\\
	If we compare the two monthly average plots for the unimodal stations, we can see that the extreme values ($\geq$ 150mm, 40 observations), has an impact on the spread in the rainiest months, but does not affect the overall pattern of the means. But it has a rather big impact on the distribution within the rainier months, especially August and September that goes from being normally distributed to being right skewed, i.e the larger half of the values have a bigger spread then the lower half. The number of outliers are also reduced by removing the extreme values. Other than that, the two plots looks very much the same, with the dry season months staying the exact same, which is to be expected.
	The semibimodal stations are less effected by removing the higher values, probably because we have so many more large observations in these stations that can compensate for a few very large values. The very obviouse difference is in June, since it becomes much more compact by removing it, but it still has one outlier. Just like for the unimodal stations, most of the outliers disappear when removing the extremes and the dry season stays invariant. The semibi stations are more normally distributed then the unimodal but about the same spread in the rainy seasons. A similar thing is observed with the bimodal stations, but we still have quite a few outliers even after removing the extremes. Only June has changed its maximum value, so removing extremes seems to make the least difference in this group. 
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveUni.pdf}
			\caption{Unimodal}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveUniN.pdf}
			\caption{Unimodal, capped}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveBi.pdf}
			\caption{Bimodal}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveBiN.pdf}
			\caption{Bimodal, capped}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveSemi.pdf}
			\caption{Semi-bimodal}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveSemiN.pdf}
			\caption{Semi bi modal, capped}
		\end{subfigure}
		\caption{Monthly averages for each rain mode group}
		\label{MonAvegroups}
	\end{figure}

\begin{table}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
\textbf{Uni} &mean &  sd& median &mad & min &  max& skew &kurtosis \\
\hline
jan  &  2.28 & 3.15 &  0.64 & 0.8 &0.02 &10.62 & 1.28 & 0.17 \\
\hline
feb  &  5.07& 5.35 &  3.12 & 3.77 &0.28 &21.43& 1.47&1.57 \\
\hline
mar  &  18.63 &8.41  &17.27 & 8.81 &5.57 &35.48& 0.23 &-0.89 \\
\hline
apr & 56.23& 13.34&  55.08  & 12.22& 29.43& 84.48 &0.21 & -0.66 \\
\hline
may  & 80.83& 16.45 &  80.72 &20.4 &44.3 &117.93 &0.04 & -0.56 \\
\hline
jun &109.71 &18.02& 107.38 &16.95 &77.48 &148.87&0.31  &-0.62 \\
\hline
jul & 114.86 &18.59 &113.22 &21.4 &87.92 &178 &  1.1 & 2.05 \\
\hline
aug &146.66& 34.63& 142.12 & 41.53& 98.38 &238.3 & 0.65  &-0.27 \\
\hline
sep &136.47& 19.23 &135.26 & 16.86 &95.57 &193.25&  0.41 &0.95 \\
\hline
oct &63.9& 15.22  &61.93 & 16.63& 33.47 &89.05& 0.06& -1.05  \\
\hline
nov &7.89 &5.2  & 6.43  & 3.25& 1.58 &20.52 &0.97 & -0.19 \\
\hline
dec& 3.04& 3.32 &  1.79  &  2.14 & 0.12& 13.38& 1.37 & 1.26 \\
\hline
\textbf{Bi} & & & & & & & & \\
\hline
jan &16.66 &6.54  & 15.6 &5.99 &4.72& 30.12& 0.26& -0.75 \\
\hline
feb  &47.25 &12.85 & 45.14 & 10.82 &23.75 &73.97&0.25 & -0.69 \\
\hline
mar &105.9 &12.98 &107.71  &11.63 &80.27& 129.5& -0.23 & -0.75 \\
\hline
apr &150.31 &18.14& 153.32 & 21.98 &117.82 &182.5 & -0.13 &-1.24 \\
\hline
may &158.76 &22.73 &158.79  &26.63 &118.87 &215.94& 0.23 &-0.23 \\
\hline
jun &188.5 &28.26 &178.88  &33.11 &147.43& 243.17 & 0.24  &-1.34 \\
\hline
jul &125.28 &26.56& 122.78 & 25.71 &83.07& 216.83 &1.19  &2.4 \\
\hline
aug &84.85 &10.92 & 85.61 &10.4 &65.28 &112.2 & 0.42 & -0.18 \\
\hline
sep &159.07& 23.79& 156.04 & 23.04 &117.25 &204.13&0.18 &-0.74 \\
\hline
oct &166.45 &17.04 &170.68  &12.21 &118.48& 196.97&-0.97 &0.89 \\
\hline
nov &62.91& 9.4 & 65.62 & 10.09 &43.55& 74.6 & -0.59& -0.92 \\
\hline
dec  &27.49 &8.21 & 27.99 & 5.16 &10.72 &52.8 &0.57 & 1.25 \\
\hline
\textbf{Semi} &  & & & & & & & \\
\hline
jan &14.71 &6.52 & 12.96 & 5.98 &5.25 &33.18& 0.93& 0.26 \\
\hline
feb &26.87 &9.73  &25.78 & 9.64 &8.12& 55.95 &0.63 & 0.74 \\
\hline
mar &61.51 &17.47 & 56.95&11.18 &35.93& 106.37 &1.04 & 0.32 \\
\hline
apr &106.34 &23.21& 103.82 & 26.96 &75.67& 162.97& 0.57 & -0.55 \\
\hline
may &181.48&26.85& 177.67 &27.71 &128.53& 249.35& 0.4 &-0.1 \\
\hline
jun &236.12 &37.12 &233.34 &34.51 &162.9& 327.5 & 0.33 & -0.13 \\
\hline
jul &78.95& 17.02 &  75.7 & 18.85& 51.75 &125.97 & 0.52& 0 \\
\hline
aug& 33.73& 8.4  &32.94 &6.3 &15.35 &50.88 & 0.08 & -0.3 \\
\hline
sep &59.62 &16.86 & 54.84 & 14.22 &35.63 &93.1 &0.53 & -0.94 \\
\hline
oct &106.59& 18.01 &109.67 & 22.68 &70.1 &135.75 &-0.19  &-1 \\
\hline
nov &55.45 &13.01 & 54.47 & 14.84 &32.93 &82.85 &0.25 & -1.03 \\
\hline
dec &28.64 &11.52 & 26.88 & 11.48 &14.2 &69.45& 1.43 & 2.85 \\
\hline
\end{tabular}
\caption{Descriptive data, monthly average}
\label{MonAveTab}
\end{table}

		\begin{figure}
			\centering
			\includegraphics[width=0.4\linewidth]{mapRainmode.pdf}
			\caption{Rain modes}
			\label{mapRainmode}
		\end{figure}
		\begin{figure}[H]
		\centering
		\begin{subfigure}{0.4\linewidth}
			\centering
			\includegraphics[width=1\linewidth]{mapObs.pdf}
			\caption{Positive observations in \%}
		\end{subfigure}%
		\begin{subfigure}{0.4\linewidth}
			\centering
			\includegraphics[width=1\linewidth]{mapObsnorm.pdf}
			\caption{Positive observation in \%, $>$ 1mm}
		\end{subfigure}
		\begin{subfigure}{0.4\linewidth}
			\centering
			\includegraphics[width=1\linewidth]{mapAnMeannorm.pdf}
			\caption{Annual mean, $>$ 1mm}
		\end{subfigure}%
		\begin{subfigure}{0.4\linewidth}
			\centering
			\includegraphics[width=1\linewidth]{mapDayMeannorm.pdf}
			\caption{Daily mean on rainy days, $>$ 1mm}
		\end{subfigure}
		\caption{Observations, annual mean, daily mean}
		\label{maps}
	\end{figure}
	Looking at the values in table \ref{MonAvegroups}, we can see that the mean and the median is very close to each other for all stations and months. The \textit{mean absolute deviation} varies a lot between months within groups and between groups, where we for example have a very low value for August in Semi but very large in Uni, which is to be expected since we have a larger spread in rainy months than dryer. We can once again see the big difference in the dryer months between the north and the other stations, with the north having a much dryer Dec-Feb then the other. Disscusion on appropriate values for the skew and kurtosis range for data to be normally distributed varies from 0.8 up to 3, with many stating 2 to be the general value. If we use 2, all months are considered to be normally symmetrical distributed but if we use 0.8 , the dry period in the north and Semi and the most extreme months in Bi is not. Nearly all months in Bi is having a negative kurtosis wheras for the other only half are negative. So many of the sets are more lightly tailed then a normal distribution.\\

	We can see that the number of positive observations varies a lot, with most observations in the midlands and the station "AXM", and fewer in the north and along rest of the coastline. We get nearly the exact same distribution whether we include rain amounts less then 1 mm or not. The annual mean is close to identical if we discard all observations $<$ 1mm, sometimes refered to as trace values, which makes sense since they are very small and will not have that much of an impact on an annual scale. But if we look at the rainy day mean, we see, as expected, an increase in the mean, but also a slightly different distribution. If we include all positive values, we have a very low mean on the station "TDI" and relatively high means in the north. But if we discard the trace values, we see a higher mean compared to the other stations along the coast and a lower mean in comparison with other stations in the north. This tells us that we have many extremly small values along the coast, but much less in the north and the midlands, which could be because of higher humidity by the ocean.

	In fig \ref{MaxAll} we can see the maximum value in each month and station. This shows that some months have a big spread and others are much closer together. We can see that the values in January and December are clustered together, except from station "NAV1" (dashed violet line) which is much lower and station "KSI" (dashed gray) in January which is much higher and "BEK" (solid blue) and "KSI" in December which both are larger. It is also clear that the spread in April and September is much smaller, and February is even closer if we ignore "NAV1". So we have a smaller spread in maximum rainfall during the rainiest months than we do in the dry season in general and a very big spread in the two transition months (March, October).

	\begin{figure}[H]
		\centering
		\includegraphics[width=1\linewidth]{MonMax.pdf}
		\caption{Maximum value in each month, all stations}
		\label{MaxAll}
	\end{figure}

	\section{Results}
	\subsection{Distribution fit}
	Because of sesonal rain behaviour, a dry period in December-February and one or two rainy seasons in April-July and September-October, each month has been considered seperatly. To also account for the different rain seasons, I have keept the three groups introduced in the previous section. By keeping the stations in groups instead of analysing each station one by one, we get a bigger data set to work with, but also automatically introduce more variation since the stations are not having identical behaviours, as seen in table \ref{Tstations} and \ref{maps}. I have decided to exclude December-February in the unimodal group because of too few data points. For all other months in all groups, I have tried and fit the data to a gamma distribution by MLE. I have thereafter plotted a histogram with the estimated denisty function, a QQ plot, the empirical and theoretical cumulitative density function and a Probability-Probability plot, to get a graphical idea if it is a resonable fit or not. In cases where the gamma distribution has not been a good fit, I have instead tried to fit the data to a lognormal, exponential and Weibull distribution. If all of them look equally good, I picked the distibution with the lowest AIC. For distribution fits where the QQ plot looks straight up to a certain value and then wonders of in some direction, I have fitted a new gamma distribution to only the values lower then that value to see if I could get a better graphical fit. If the QQ plot looked better after refitting the data, I used the goodness-of-fit tests Kolmogorov-Smirnov, Anderson-Darling and Cramer-von-Misen, to test if they are in fact better fits. 
	\subsubsection{Unimodal stations (KRA, BOL, TLE, NAV1)}
	As many other has discovered, the rain distribution in the north only has one rainy season, which clearly can be seen in fig \ref{MonAvegroups}. The only missing data points in this group is from one full November and one full December and it has got proportionaly equally many positiv observations as the semi-bimodal group, but very differently distributed. By looking at QQ plots of the months (fig \ref{GammaFitUni}), we can see that June and August fits perfectly well to their distributions, and May only has a couple of observations that does not fit. But April, July, September and October fits poorly after 80 mm and March after 60 mm. November is a poor fit in general, but that most likely depends on the lack of data and can therefore not be improved. 
	\begin{figure}[H]
		\ContinuedFloat*
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitMar.jpg}
			\caption{March,(15.633, 0.633) }
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitApr.jpg}
			\caption{April, (19.673, 0.674)}
		\end{subfigure}
		\caption{Gamma distribution fits, Unimodal group}
		\label{GammaFitUni}
	\end{figure}
	\begin{figure}[H]
		\ContinuedFloat
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitMay.jpg}
			\caption{May, (18.326, 0.705)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitJun.jpg}
			\caption{June, (19.902, 0.703) }
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitJul.jpg}
			\caption{July, (19.678, 0.663)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitAug.jpg}
			\caption{August, (22.687, 0.614)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitSep.jpg}
			\caption{September, (17.848, 0.716)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitOct.jpg}
			\caption{October, (15.229, 0.726)}
		\end{subfigure}
		\caption{Gamma distribution fits, unimodal group}
	\end{figure}
	To try and improve the fit, one can cut of all values larger than where they start to wander off and fit a seperate extreme distribution to the higher values. By trying different splitting values, it was possible to improve the fit, at least graphically, for all months except April (fig \ref{UniCap}). But when tested with the three mentioned goodness-of-fit tests, it showed that only some of the months got a better fit with the capped data set. Both October and March got much higher p-values (March: KS 6\% to 10\%, CVM 3\% to 6\%, October: KS 4\% to 6\%, CVM 3\% to 4\%) with the capped data sets with all the tests, April and July showed no improvment and September even performed worse. The biggest change in the scale value is seen in March, which could potentially be since it is the transition month into the rainy season, but the scale value changed a lot less for October, which is the other transition month. 
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.4\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitMar70.jpg}
			\caption{March, Data $<$ 70, (12.805, 0.681)}
		\end{subfigure}%
		\begin{subfigure}{0.4\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitJul120.jpg}
			\caption{July, Data $<$ 120, (18.668, 0.677)}
		\end{subfigure}
			\begin{subfigure}{0.4\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitSep100.jpg}
			\caption{September, Data $<$ 100, (16.5, 0.74)}
		\end{subfigure}%
			\begin{subfigure}{0.4\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitOct100.jpg}
			\caption{October, Data $<$ 100, (14.433, 0.744)}
		\end{subfigure}
		\caption{Gamma fits with capped data}
		\label{UniCap}
	\end{figure}	
	\subsubsection{Bimodal stations (ODA, KDA, KSI, HO, SUN, WEN)}
	For these stations, we see a very similar pattern in the montly distributions as for the unimodal group. Just like there, we have that March is right skewed i.e the data has more large values then expected from the fitted distribution and April left skewed i.e less large values then expected, May is a good fit except for the few largest values and September gets a poorer fit after 70 mm. But here June and August are much more left skewed then for the unimodal stations and July appears not to be gamma distributed at all. The dry season months show a nearly perfect fit. 
		\begin{figure}[H]
		\ContinuedFloat*
		\centering
		\begin{subfigure}{0.4\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitJan.pdf}
			\caption{January, (15.983, 0.648)}
		\end{subfigure}%
		\begin{subfigure}{0.4\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitFeb.pdf}
			\caption{February, (15.275, 0.705)}
		\end{subfigure}
		\caption{Gamma distribution fits, bimodal group}
		\label{GammaFitBi}
	\end{figure}
		\begin{figure}[H]
		\ContinuedFloat
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitMar.pdf}
			\caption{March,(18.644, 0.666)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitApr.pdf}
			\caption{April, (19.447, 0.737)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitMay.pdf}
			\caption{May, (18.246, 0.691)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitJun.pdf}
			\caption{June, (20.498, 0.627)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitJul.pdf}
			\caption{July, (19.122, 0.532)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitAug.pdf}
			\caption{August, (13.068, 0.547)}
		\end{subfigure}
		\caption{Gamma distribution fits, bimodal group}
	\end{figure}
	\begin{figure}[H]
		\ContinuedFloat
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitSep.pdf}
			\caption{September, (16.732, 0.625)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitOct.pdf}
			\caption{October, (14.17, 0.73)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\includegraphics[width=1\linewidth]{BiFitNov.pdf}
			\caption{November, (11.099, 0.736)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\includegraphics[width=1\linewidth]{BiFitDec.pdf}
			\caption{December, (14.223, 0.672)}
		\end{subfigure}
		\caption{Gamma distribution fits, bimodal group}
	\end{figure}
	Since July and August are so skewed, I tried to fit them to some other distributions and found that they fitted much better to a Weibull distribution. Fot the other ones, I kept the gamma distribution and only changed the splitting value. Once again, all of the months graphically improves a lot by this method but only a few get larger p-values in the goodness-of-fit tests. July and August improves greatly (July: CVM $3.1e^{-12}$ to $6.1e^{-7}$, KS 0 to $5.8e^{-9}$, August: CVM $2.2e^{-11}$ to $2.3e^{-8}$, KS 0 to $4.8e^{-12}$) by fitting them to a Weibull distribution instead of a gamma, June and September improves slightly (CVM: June $1.56e^{-5}$ to $1.77e^{-5}$, September $3.8e^{-7}$ to $7e^{-7}$) by adding the cap, but October shows no improvment and March performs worse. So here we see an improvment in the rainy months and no improvment in the transition months, right opposite to results for the unimodal stations.
	\begin{figure}[H]
		\ContinuedFloat*
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{FitBiMar80.pdf}
			\caption{March, Data $<$ 80, (17.446, 0.683)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{FitBiJun150.pdf}
			\caption{June, Data $<$ 150, (20.145, 0.631)}
		\end{subfigure}
	\caption{Distribution fits with capped data}
	\label{BiCap}
	\end{figure}
	\begin{figure}[H]
		\ContinuedFloat
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{FitBiJulWei.pdf}
			\caption{July, Weibull distribution, (6.976 0.661)}
			\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{FitBiAugWei.pdf}
			\caption{August, Weibull distribution, (5.014, 0.666)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{FitBiSep90.pdf}
			\caption{September, Data $<$ 90, (15.73, 0.64)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{FitBiOct90.pdf}
			\caption{October, Data $<$ 90, (13.658, 0.741)}
		\end{subfigure}
		\caption{Distribution fits with capped data}
		\end{figure}

	\subsection{Semi-bimodal stations (AXM, TDI, SAL, ACC, ADA, TEM)}
	These stations do not seem to come from a gamma distribution since I get a poor fit for nearly all months except March and December, if we discard the largest value. January, February, May, June and November are well fitted for most of the values, but July, August, September and October diverges from the straight line very quickly. So it seems like the dry season fits much better then the rainy season.
	As I put a cap on June data, it became clearer that the lower part was very left skewed, i.e that the data contains many more points in the mid range then the fitted distribution, and that the data fitted better to a Weibull distribution, eventhough it is far from optimal. Regardless of distribution and cap, I see no improvment in the distribution fit on July or August. One possible way to deal with this is to divide the data into low, moderate and heavy rainfall, and fit different distributions to each set. The general behaviour of the months is that we have fewer observations in the low-mid range than expected and more higher values than expected. We can improve the graphical fit in January, February, April and May by capping the maximum value and June by fitting it to a Weibull distribution instead. But the other months has the pattern described above and does therfore not become any better by just adding caps or changing to a Weibull or lognormal distribution.

	\begin{figure}[H]
		\ContinuedFloat*
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitJan.pdf}
			\caption{January, (15.346, 0.599)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitFeb.pdf}
			\caption{February, (17.008, 0.583)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitMar.pdf}
			\caption{March,(19.727, 0.627)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitApr.pdf}
			\caption{April, (22.747, 0.588)}
		\end{subfigure}
	\caption{Gamma distribution fits, semi-bimodal group}
	\label{GammaFitSemi}
	\end{figure}
	\begin{figure}[H]
		\ContinuedFloat
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitMay.pdf}
			\caption{May, (24.726, 0.591)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitJun.pdf}
			\caption{June, (30.324, 0.529)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitJul.pdf}
			\caption{July, (21.233, 0.465)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitAug.pdf}
			\caption{August, (7.705, 0.564)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitSep.pdf}
			\caption{September, (11.671, 0.519)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitOct.pdf}
			\caption{October, (19.981, 0.538)}
		\end{subfigure}
		\caption{Gamma distribution fits, semi-bimodal group}
	\end{figure}
		\begin{figure}[H]
			\ContinuedFloat
			\centering
		\begin{subfigure}{0.5\textwidth}
			\includegraphics[width=1\linewidth]{SemiFitNov.pdf}
			\caption{November, (15.698, 0.621)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\includegraphics[width=1\linewidth]{SemiFitDec.pdf}
			\caption{December, (16.805, 0.575)}
		\end{subfigure}
		\caption{Gamma distribution fits, semi-bimodal group}
	\end{figure}
	By adding a cap on January we get a better score on the CVM test but worse on the KS test. AD gives the same pvalue. February improves with a power of 2 on all tests by adding the cap. By changing the cap from 130 to 90 we get a better KS value but all the other stays the same, if I use 130 as a cap, the tests performs worse than for the complete data set. May behaves like January, I get better scores with the AD and CVM test but worse with KS. Just like for the bimodal stations, I get a huge improvement by fitting June data to a Weibull insted of a gamma(CVM: $1.3e^{-10}$ to $2.5e^{-5}$, KS: $2.8e^{-14}$ to $3.6e^{-6}$). 
		\begin{figure}[H]
			\ContinuedFloat*
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitJan70.pdf}
			\caption{January, Data $<$ 70, (13.844, 0.622)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitFeb70.pdf}
			\caption{February, Data $<$ 70, (14.616, 0.612)}
		\end{subfigure}
		\caption{Distribution fits with capped data}
		\label{SemiCap}
	\end{figure}
	\begin{figure}[H]
		\ContinuedFloat
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitApr120.pdf}
			\caption{April, Data $<$ 120, (21.738, 0.598)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitMay130.pdf}
			\caption{May, Data $<$ 130, (23.325, 0.603)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitJunWei.pdf}
			\caption{June, Weibull, (11.451, 0.646)}
		\end{subfigure}
		\caption{Distribution fits with capped data}
		\label{SemiCap}
	\end{figure}

\section{Comparing data to CMIP5}
\subsection{Rain over a threshold}
Studying figure \ref{PRCPTOT}, \ref{R1mm}, \ref{R10mm} and \ref{R20mm} we can see a very clear pattern in the difference between our data and the GCMs (Global Climate Models). The GCMs heavily overestimates the number of rainy days which also leads to a vast overestimation of the annual total rainfall (figure \ref{PRCPTOT}, \ref{R1mm}). The span among the GCMs is also very large, ranging from 600 mm up to 2800 mm per year, wheras the range when splitting the data into the same rain modes as before, is only between 500 and 1500 mm per year. So it appears that at least a few of the models simulates in the correct range, but looking at the CMIP5 mean it is evient that the majority of the GCMs simulates  too much annual rainfall. However, the behaviour of the data curve and the CMIP5 mean curve is similar, which could mean that the GCMs can correctly simulate the changes between years even if they cannot correctly simulate the number of rainy days.

Looking at figure \ref{R1mm}, we can see that all models simulate more rainy days than our data, which leads to the mean curve to be about 100 days per year shifted compare to the data curve. But the range between the different GCMs is once again large, ranging from 100-300 days. Looking at the CMIP5 mean curve, we can see that the models are not as good at simulating the variation in rainy days between years as they are at simulating changes in rain amount, since the plot is relatively flat.

When looking at heavier rainfall, the second known issue becomes clear. The GCMs can simulate the number of days with $\geq$ 10 mm fairly well, the curve is slightly higher than our data but not completly out of range, wheras for $\geq$ 20 mm, they simulate too few days. This is a well know problem, the oversimulation of rainy days and the lack of skill to simulate days with very heavy rainfall. Another interesting thing to notice is that the CMIP5 mean curve seems to have a uppward pointing trend for both $\geq$ 10mm and $\geq$ 20 mm which is not clear in our data. This could be an issue if we want to use these models to predict future behaviour of precipitation.

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{PRCPTOTall.pdf}
		\caption{All stations}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{PRCPTOTmodes.pdf}
		\caption{Modes}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{PRCPTOTGCMall.pdf}
		\caption{CMIP5 all}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{PRCPTOTGCMmean.pdf}
		\caption{CMIP5 mean}
	\end{subfigure}
	\caption{Total annual precipitation ($\geq$1 mm)}
	\label{PRCPTOT}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R1mmall.pdf}
		\caption{All stations}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R1mmodes.pdf}
		\caption{Modes}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R1mmGCMall.pdf}
		\caption{CMIP5 all}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R1mmGCMmean.pdf}
		\caption{CMIP5 mean}
	\end{subfigure}
	\caption{Number of rainy days ($\geq$1 mm)}
	\label{R1mm}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R10mmall.pdf}
		\caption{All stations}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R10mmodes.pdf}
		\caption{Modes}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R10GCMall.pdf}
		\caption{CMIP5 all}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R10GCMmean.pdf}
		\caption{CMIP5 mean}
	\end{subfigure}
	\caption{Number of rainy days ($\geq$10 mm)}
	\label{R10mm}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R20mmall.pdf}
		\caption{All stations}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R20mmodes.pdf}
		\caption{Modes}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R20GCMall.pdf}
		\caption{CMIP5 all}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R20GCMmean.pdf}
		\caption{CMIP5 mean}
	\end{subfigure}
	\caption{Number of rainy days ($\geq$20 mm)}
	\label{R20mm}
\end{figure}

\subsection{Rain quantiles}
Since CMIP5 uses earlier years then what we have data from as reference period, it is not possible to compare the numbers with each other, but more the behaviour of the curve. 84-93 is picked as a reference period instead of 83-92 to avoid the clearly lower values in -83. In figure \ref{R95} both curves seems to exhibit a very similar behaviour, which is a steady increase in the total rainfall on days with heavy rainfall. For days with extreme rainfall (figure \ref{R99}), the simulated mean is very close to the data mean, but the spread among the models is still very large. The CMIP5 mean is again showing a steady inscrease which is not clearly visible in the data plot. So similar differences can been seen both when looking at the highest percentiles and very heavy rainfall in mm. 
\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R95all.pdf}
		\caption{All stations}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R95GCMall.pdf}
		\caption{CMIP5 all}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R95GCMmean.pdf}
		\caption{CMIP5 mean}
	\end{subfigure}
	\caption{Total rain amount in days above 95\% threshold for reference period}
	\label{R95}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R99all.pdf}
		\caption{All stations}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R99GCMall.pdf}
		\caption{CMIP5 all}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R99GCMmean.pdf}
		\caption{CMIP5 mean}
	\end{subfigure}
	\caption{Total rain amount in days above 99\% threshold for reference period}
	\label{R99}
\end{figure}

\subsection{Future predictions CMIP5}
One of the major reasons to build GCMs is to be able to simulate future changes in the climate. However, since the models can not exactly simulate the climate that we have observed, one can debate how accurate the predictions are. But in general they can somewhat accurately simulate changes due to changes in the ocean and atmospheric dynamics, even if they still cannot predict correct changes in absolute numbers. Another resaon to not fully trust the GCMs future predictions, is the spread between their simulations. If we look at fig \ref{R1GCM45}, we can see that there are two seperate groups, with the lower group more accurately simulation our observed data. The group with more rainy days seems to exhibit a decreasing trend in the future, which heavliy influences the CMIP5 mean curve, but the lower group looks pretty stationary. Should we put more weight on the simulations that more accurately mimic our data, and therefor not assume that number of rainy days will decrease, or should we put equal weight to all models and believe that number of days will decrease just from a lower value then what the CMIP5 thinks?

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{R1GCM45all.pdf}
		\caption{All models}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{R1GCM45mean.pdf}
		\caption{Mean}
	\end{subfigure}
	\caption{R1mm for rcp45 scenario 1980-2100}
	\label{R1GCM45}
\end{figure}
Studying fig \ref{R10GCM45} makes it more difficult to talk about increase in extreme weather, since it has been a decline from 1850 until around 1980, and now it is predicted to rise to the same level as 1850. So one can discusse if we are going to see more extreme weather, or if we are experiencing extreme weather now and we are on the way back to a normal state. If one were just to look at the mena plot form 1980 and forward it looks like we are only experiencing a steady increase away from what we probably would assume is normality. Figure \ref{R20GCM45} also shows an increase, but with around only one day over a 100 year period. So from a per cent point of view, it is a large increase, but does it have an impact in reality? And once again, even if the mean is showing an increase, if we study all of the models, there is still a massive range both in the beginning and the end of the period with some models still simulating 0 days. 
\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R10GCM45all.pdf}
		\caption{All models}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{R10GCM45mean_long.pdf}
		\caption{Mean}
	\end{subfigure}
	\caption{R10mm for rcp45 scenario 1980-2100}
	\label{R10GCM45}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{R20GCM45all.pdf}
		\caption{All models}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{R20GCM45mean.pdf}
		\caption{Mean}
	\end{subfigure}
	\caption{R20mm for rcp45 scenario 1980-2100}
	\label{R20GCM45}
\end{figure}

\section{Trends in timeseries, Lowess}
Studying figure \ref{lowess}, it is quite evident that the annual precipitation, very heavy rainfall and the rain amount on R95 (days with more rain then the 95 percentail in the reference period), is increasing over the 30 year period. Number of rainy days is showing a small increase aswell. Number of days with $\geq$ 10 mm is not showing a consistent pattern since it is a small decrease for the first half of the period, to then increase back to the same level as 1983. R99 is showing the opposite pattern, an increase in the amount for the first half of the period to then decrease back to the 1983 level. 
This could suggest that there is an increase in the number of rainy days and an increase in days with very heavy rain but a slight decrese in the most extreme rainfalls. 
\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{lowessAp.pdf}
		\caption{Annual precipitation}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{lowessRd.pdf}
		\caption{Number of rainy days}
	\end{subfigure}	
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{lowessR10.pdf}
		\caption{Days $\geq$ 10mm}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{lowessR20.pdf}
		\caption{Days $\geq$ 20mm}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{lowessR95.pdf}
		\caption{Days $\geq$ 95\%}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{lowessR99.pdf}
		\caption{Days $\geq$ 99\%}
	\end{subfigure}
	\caption{Smooting using lowess}
	\label{lowess}
\end{figure}

If we leave the changes in extreme indicies and instead look at some extreme values for each year, figure \ref{extRainRatio} shows that the ratio of very heavy rainfall ($\geq$ 95\% quantile) is more or less constant with large flucations around -98. The extreme rainfall ratio is instead showing a downward trend. Since figure \ref{lowess} (a) show that annual precipitation is increasing over time, this indicated that the extreme rainfall is either constant or not increasing in the same tempo as the total annual rainfall. Autocorrelation plot shows no significant correlation for either time series. 
\begin{figure}[H]
	\centering
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{extRainRatio95.pdf}
		\caption{95\% quantile}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{extRainRatio99.pdf}
		\caption{99\% quantile}
	\end{subfigure}	
	\caption{Ratio extreme rainfall to total annual rainfall}
	\label{extRainRatio}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsPRCPTOTu.pdf}
		\caption{Uni}
	\end{subfigure}%
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsPRCPTOTb.pdf}
		\caption{Bi}
	\end{subfigure}%
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsPRCPTOTs.pdf}
		\caption{Semi}
	\end{subfigure}
	\caption{Total annual precipitation}
	\label{modesPRCPTOT}
\end{figure}

\subsection{Rain modes}
\begin{figure}[H]
	\centering
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR1u.pdf}
		\caption{Uni}
	\end{subfigure}%
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR1b.pdf}
		\caption{Bi}
	\end{subfigure}%
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR1s.pdf}
		\caption{Semi}
	\end{subfigure}
	\caption{Total number of rainy days}
	\label{modesR1}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR10u.pdf}
		\caption{Uni}
	\end{subfigure}%
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR10b.pdf}
		\caption{Bi}
	\end{subfigure}%
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR10s.pdf}
		\caption{Semi}
	\end{subfigure}
	\caption{Total number of days $\geq$ 10 mm}
	\label{modesR10}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR20u.pdf}
		\caption{Uni}
	\end{subfigure}%
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR20b.pdf}
		\caption{Bi}
	\end{subfigure}%
	\begin{subfigure}{0.3\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{tsR20s.pdf}
		\caption{Semi}
	\end{subfigure}
	\caption{Total number of days $\geq$ 20 mm}
	\label{modesR20}
	\end{figure}

	\subsection{Autocorrelation}
	For the threshold based inidicies there is a weak neagative autocorrelation between consecutive years, with R10 being close to significant. For the quantile based indicies there is instead a positive autocorrelation with a 6 year lag (el ni\~{n}o?).

	\section{Summary and conclusions}
	Fitting daily rainfall data is of interest in all parts of the world sice it has a great impact on life from local to global scale. Several different methods and distributions has been proposed, and fitted with different level of success. Since the rain patterns differ between countries and seasons, different distributions are suitable in each case. To give more flexibility to the distribution, it has been proposed to mix two distribution from the same family, but with different parameters and then weight them with a probability parameter, so one part models the bulk of the data and the other one the tail, but without cutting of the data at some value. Two common practices to model the occurence of rainfall is either by a binomial distribution or by a Markov chain. In this report, I have looked at if it is possible to improve the distribution over daily rainfall amount by ignoring values that departs in the QQ plot, when the full data set is fitted. By plotting new QQ plots, I could conclude that we can get a much better graphical fit by doing this, but that the goodness-of-fit test only returned better p-values for some of the refitted distributions. For the unimodal group, I got a better fit for the transition months but a worse or unchanged p-value for the rainy months. For the bimodal group I instead got an improved fit for the rainier months and no improvment in the transition months. The greates improvment was obtained by fitting July and August to a Weibull distribution instead. Lastly I got an improved or unchanged fit for all months in the semimodal group by adding a capping value and once again a great improvment by fitting June to a Weibull distribution. The reason for the semi group to be improved in more months then the other groups could be that we have more extreme values in this region. 
	\subsection{Further work}
	More work needs to be done in finding a system for when we get an improved fit in the distributions by excluding the largest values and how such a cap effectively can be found. By doing this, rainfall predictions could be improved in general and not just for this data set. Focus will then be on the extreme part of the data set, mainly by looking at what the current risk of extreme rainfall events is and how that has changed over time due to climate change. This will involve using several tools in extreme value theory. By applying the block maxima method on annual and monthly blocks, one can look at the distribution of these maximas, and get a probability distribution for these extreme events from a small sample. By instead using the peak over threshold method, we can get a distribution to infer the risk of events over a certain value. One can than start to answer questions such as, "what is the return period of a certain extreme event" and "has the probability of this event changed due to climate change". This would need to be done on a monthly basis because of the seasonal behaviour. Once I have these distributions, I can then start to compare the distributions with the representation in TAMSATv3 of these extremes. By doing this, we can start to understand how accurately TAMSATv3 is estimating extremes, to later improve this. This can then be used to investigate the relation between extreme rainfall and agriculture risk and how this is affected by changes in the extreme behaviour. To do this, I will use crop models already used in other TAMSAT projects, so they have therfore already been tested on real data.

\end{document}

### 2018-05-18 WorkingReport.tex
first upload

\documentclass{article}
\usepackage{latexsym, amssymb, amsmath, amsfonts, bbm, graphicx,float, textcomp}
\usepackage{caption}
\DeclareCaptionLabelFormat{cont}{#1~#2\alph{ContinuedFloat}}
\captionsetup[ContinuedFloat]{labelformat=cont}
\usepackage{subcaption}
\usepackage[margin=2.5cm]{geometry}
\author{Jennifer Israelsson}

\begin{document}
	\title{Working report}
	\maketitle

	\section{Preliminaries}

	\subsection{Distributions}
	\textbf{Gamma distribution: } The gamma distribution is a two-parameter continuous distribution family, characterized by a shape parameter, $\alpha$, and a scale parameter, $\theta$, both positive. It has got PDF, mean and variance,
	\begin{equation}
	f(x ; \alpha, \theta) = \frac{x^{k-1} e^{-\frac{x}{\theta}}}{\theta^k \Gamma(k)}
	\end{equation}
	\begin{center}
	$E(X) = \alpha\theta$, $Var[X] = \alpha\theta^2$
	\end{center}

	\subsection{Statistical test}
	\textbf{Wilcoxon signed rank test:} Is a non-parametric test that either tests the null hypothesis that a sample has a certain mean or the null hypothesis that two paired samples comes from the same distribution. In the paired case, you take the difference between each pair, rank them from smallest to largest absolut value and then and then add up all the ranks of the positive differences. If this value is greater then the table value for our number of pairs, we reject the null hypothesis and conclude that the data comes from different distributions.\cite{PennState} \linebreak
	\textbf{Autocorrelation plot:} Autocorrelation describes how correlated the time series data is to itself, i.e how dependent is the data a few days forward on the day today. By plotting correlation as a function on number of lag days, we can see roughly how long a wet spell is in each month. For autocorrelation to work, there can be no missing values, so one needs to remove the months from the data with missing values. 
	\subsection{Extreme value theory}
		\subsubsection{Maximum analysis}
		In extreme value theory, we only look at the larger values in our data instead of all the data, to get a better understanding of how the tail behaves, and gives us a more accurate way to extrapolate outside our data. Put $M_n$ = max$(X_1,...,X_n)$, i.e the maximum value of our n sample points. This is often called a block maxima. Lets also denote $x_F$ = sup\{$x\in\mathbb{R} ; F(x)<1$\}, the right end point of our distribution F. We are interested in finding the distribution of $M_n$ as n increases. We call a non-degenerated rv X \textit{max-stable} if
		\begin{equation}
		c_nX + d_n = M_n
		\end{equation}
		for all n $\geq$ 2 and appropriate $c_n>0$, $d_n \in \mathbb{R}$. This can also be written as
		\begin{equation}
		\mathbb{P}(M_n \leq x) = F^n(x)
		\end{equation}
		with F being the distribution function of the rv X. This leads us to the important \textit{Fisher-Tippet theorem}, which states that; if there exists norming constants $c_n$ and $d_n$ as above and some non-degenerated distribution function H such that
		\begin{equation}
		c_n^{-1}(M_n - d_n) \xrightarrow{d} H
		\end{equation}
		then H belongs to one of the three extreme value distributions; Fr\'{e}chet, Gumbel and Weibull. These can be summarised in one distribution, called \textit{Generalised extreme value} distribution, GEV. 
		\begin{equation}
		H_\xi(x) = 
			\begin{cases}
				\text{exp}(-(1+\xi x)^{-1/\xi}), & \text{if} \ \xi \neq 0 \\
				\text{exp}(-(\text{exp}(-x))), & \text{if} \ \xi = 0 
		\end{cases}
		\end{equation}
		where $1+\xi x > 0$. With $\xi = 0$, we get Gumbel, $\xi > 0$ Fr\'{e}chet and $\xi < 0$ Weibull. Closely related to GEV distributions is \textit{maximum domain of attraction}. We say that a rv X belongs to the \textit{maximum domain of attraction} of \textit{H}, X $\in MDA(H)$, if there exists norming constants $c_n$ $>$ 0, $d_n \in \mathbb{R}$ such that
		\begin{equation}
			c_n^{-1}(M_n - d_n) \xrightarrow{d} H
		\end{equation}

		\subsubsection{Upper order analysis}
		Instead of looking at maxima, and thereby throwing away a lot of data, we can look at all values above a threshold, u. We can look at both the distribution and the mean of the data above this threshold
		\begin{equation}
		F_u(x) = \mathbb{P}(X - u \leq x | X > u) = \frac{F(x + u) - F(u)}{1 - F(u)}
		\end{equation}
		\begin{equation}
		e(u) = \mathbb{E}(X - u | X > u)
		\end{equation}
		$e(u)$ is called \textit{the mean excess function of X}. Just like the block maxima converges to a GEV distribution, the data over a high threshold also converges to a distribution, called the \textit{General pareto distribution}, GPD
		\begin{equation}
			G_{\xi, \nu, \sigma}(x) =
				\begin{cases}
				1-(1+\xi \frac{x-\nu}{\sigma})^{-1/\xi}, & \text{if} \ 1+\xi \frac{x-\nu}{\sigma} > 0, \xi \neq 0 \\
				1- e^{-\frac{x-\nu}{\sigma}}, & \text{if} \  \xi = 0
				\end{cases}
		\end{equation}
		where
		\begin{equation}
			\begin{cases}
				x \geq 0, & \text{if} \ \xi \geq 0 \\
				0 \leq x \leq -\sigma/ \xi, & \text{if} \ \xi < 0
		\end{cases}
		\end{equation}
		$\nu \in \mathbb{R}$ is called a location parameter and $\sigma>0$ is a scale parameter. The two distributions are related, such that the distributions for which the block maxima converges to a GEV with parameter $\xi$, their excess distribution converges to the GPD with same shape parameter $\xi$.

	\section{Initial assesment}
	To get an idea about the general behaviour of the data, and if there are big variations depending on the geography, I slit all the data into the different stations. To get a solid baseline to compare the rest of the data to, I picked out 3 complete data sets; one in the north (ADA), one around the equator (KRA) and one in the south (BOL). As a first attempt to see differences, I picked out only the rainy days from the data and compared the number of days. I then tried to fit this data into a gamma distribution by using Maximum likelihood estimator (MLE). To get a bit deeper understanding of their differences, I split the positiv data into months and looked at the behvaiour in the months; does it have a few extreme values, many or few days of rain, certain years with many days of rain (rainy) or several days with a high amount of rain (wet). As a final thing, I picked out the maximum values in each month and the years for all the extreme values.

	\subsection{ADA}
	This station has a lot fewer days of rain compared to the other two station, with only 19\% of days rainy, but it is more evenly distributed then for the other stations. The number of days of rain for (Dec, Jan, Feb) is (37, 21, 58), which is higher then for the other. It fits to a \textbf{Gamma distriution} with \textbf{scale}=21.02 and \textbf{shape}=0.553 and has a loglikelihood of -6730. This distribution fits nicely up until 100 quantile (fig \ref{GammaADA}), so instead I only fitted the data $<$ 100 and got a \textbf{Gamma distribution} with \textbf{scale} = 17.590 and \textbf{shape} = 0.585 (fig \ref{GammaADAn100})
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{QQADAfull.png}
			\caption{Full data set}
			\label{GammaADA}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{QQADAn100.png}
			\caption{Positive values $<$ 100}
			\label{GammaADAn100}
		\end{subfigure}
		\caption{Q-Q plots for Gamma distribution, ADA}
	\end{figure}
 	\subsection{BOL}
 	This station has approxiamtely 24\% of rainy days with very few in (Dec, Jan, Feb), (7,29,21). The \textbf{Gamma distribution} with \textbf{scale}=17.835, \textbf{shape}=0.689 fits well up to 80-100 quantile if fitted with all the data (fig \ref{GammaBOL}), and if I only fit it with values up to 100, I get a \textbf{Gamma distribution} with \textbf{scale} = 17.249, \textbf{shape} = 0.698 which fits nicely nearly everywhere (fig \ref{GammaBOLn100}).
 	\begin{figure}[H]
 		\centering
 		\begin{subfigure}{0.5\textwidth}
 			\centering
 			\includegraphics[width=1\linewidth]{QQBOLfull.png}
 			\caption{Full data set}
 			\label{GammaBOL}
 		\end{subfigure}%
 		\begin{subfigure}{0.5\textwidth}
 			\centering
 			\includegraphics[width=1\linewidth]{QQBOLn100.png}
 			\caption{Positive values $<$ 100}
 			\label{GammaBOLn100}
 		\end{subfigure}
 		\caption{Q-Q plots for Gamma distribution, BOL}
 	\end{figure}
	\subsection{KRA}
	This is the rainiest station out of the three, with $\sim$ 27\% rainy days, but less well distributed then the ADA. In (Dec, Jan, Feb), it rained (20,35, 30), so values between the other two. The \textbf{Gamma distribution} with \textbf{scale} = 23.038 and \textbf{shape} = 0.595 fits up to around 100 quantile (fig \ref{GammaKRA}) and if we restrict the data to $<$ 120, it fits nearly perfectly to a \textbf{Gamma distribution} with \textbf{scale} = 21.774 and \textbf{shape} = 0.607 (fig \ref{GammaKRAn120})
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{QQKRAfull.png}
			\caption{Full data set}
			\label{GammaKRA}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{QQKRAn120.png}
			\caption{Positive values $<$ 120}
			\label{GammaKRAn120}
		\end{subfigure}
		\caption{Q-Q plots for Gamma distribution, KRA}
	\end{figure}
 	\subsection{Max values and Extreme years}
 	\begin{table}[H]
 		\centering
 		\begin{tabular}{|c| c c c c c c c c c c c c|} 
 			\hline
 			 & Jan & Feb & Mar & Apr & May & Jun & Jul & Aug & Sep & Oct & Nov & Dec \\ 
 			\hline\hline
 			ADA & 31.3 & 74.4 & 82.6 & 124.9 & 80.8 &  205.7 & 93 & 53.6& 102.1 & 100.3 & 98.4 & 65 \\ 
 			\hline
 			KRA & 46.3 & 52.6 & 68.4 & 94.4 & 165.4 & 149 & 150.8 & 152.2 & 168.6 & 180.7 & 65.8 & 51.9 \\
 			\hline
 			BOL & 37.1 & 79.8 & 110.3 & 81.3 & 92.6 & 101.1 & 130.1 & 122 & 139.2 & 94.7 & 74 & 45.3 \\
 			\hline
 		\end{tabular}
 		\caption{Table of yearly maximum}
 		\label{tableymax}
 	\end{table}

 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{Mmaxplot.png}
 		\caption{Max values per month over all years}
 		\label{Mmax}
 	\end{figure}
 	\section{Extreme fitting}
 	To investigate the extreme bahaviour of the different stations, I started to look at the distribution of the yearly maxima, and ploted it to see if it had an overall behaviour. To determine a good threshold value, I tried to look at \textbf{"Hill plots"} and \textbf{"Mean excess plots"}, but neither gave a very clear indication for a suitable value. So instead, I looked at threshold parameter plots, starting with a range (10, 80), but moved down to (10, 60) (fig \ref{T1060} because the varinace of the higher values where too big, and therefore draged out the scale too much. 
 	\subsection{Threshrange plots}
 	\begin{figure}[H]
 		\centering
 		\begin{subfigure}{0.5\textwidth}
 			\centering
 			\includegraphics[width=1\linewidth]{ThresADA.jpg}
 			\caption{ADA}
 			\label{TADA}
 		\end{subfigure}%
 		\begin{subfigure}{0.5\textwidth}
 			\centering
 			\includegraphics[width=1\linewidth]{ThresKRA.jpg}
 			\caption{KRA}
 			\label{TKRA}
 		\end{subfigure}
 		\begin{subfigure}{0.5\textwidth}
 			\centering
 			\includegraphics[width=1\linewidth]{ThresBOL.jpg}
 			\caption{BOL}
 			\label{TBOL}
 		\end{subfigure}
 		\caption{Threshold range =(10, 60) plots}
 		\label{T1060}
 	\end{figure}

	All of the stations have very similar threshold plots, the variance starts to increase substantially around 30 and is large around 50. For ADA, the variance is already quite large at 40, wheres for the other two it is more around 45 it start to grow rapidly. So suitable thresholds could be 35 for ADA and 40 or 45 for the other 2.
 	\subsection{ADA}
 	The annual maximum for ADA has a rough pattern of getting a new maximum value every 5-7 years until 2005 (fig \ref{ADAyMax}), but it could of course be a new maximum in 2013, which would then continue the pattern. The values fits a \textbf{GPD} with parameters (69.649, 23.268, 0.088) relatively well, with a small bump on the density plot around 150, which is present for all the stations (fig \ref{ADAyMfit}). The fitted distribution understimates the rain amount for higher return years.
 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{ADAyMaxPlot.png}
 		\caption{Annual maxima, ADA}
 		\label{ADAyMax}
 	\end{figure}
 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{ADAyMaxfit.jpg}
 		\caption{Plots over distribution fit for ADA anuual max}
 		\label{ADAyMfit}
 	\end{figure}
 	If I tried to fit the data with the threshold set to 40 (140 obs.), only the lower values are well fit, and not the tail which is the one of interest. So instead I picked the threshold to be 50 (84 obs.)to better fit the tail (fig \ref{T50ADA}). It is a pretty big bump around 100 in the density plot and a small one around 150. The fitted distribution is with parameters (60.806, 10.714, 0.407). 
 		\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{T50ADA.jpg}
 		\caption{Plots with threshold = 50, ADA}
 		\label{T50ADA}
 	\end{figure}

 	I plot the data, highlighting only the data points that exceeds the threshold, to see if there is a general trend with more extreme values (fig \ref{ADAPOT}). It seems like there where more days with more then 50 mm of rain before 2000 then after.
 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{ADAPOTplot.png}
 		\caption{Plots with threshold = 50, ADA}
 		\label{ADAPOT}
 	\end{figure}
 	\subsection{BOL}
 	The annual maxima for BOL seems to have a two-peak periodic pattern (El nino?) (fig \ref{BOLyMax}), and the maximum values fits fairly good to a \textbf{GPD} with parameters (70.312, 16.860, 0.147) (fig \ref{BOLyMfit})., but with a big bump around 130. This distribution matches the return levels much better then for ADA. 		\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{BOLyMaxPlot.png}
 		\caption{Annual maxima, BOL}
 		\label{BOLyMax}
 	\end{figure}
 		\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{BOLyMaxfit.jpg}
 		\caption{Plots over distribution fit for BOL anuual max}
 		\label{BOLyMfit}
 	\end{figure}
 	If I use 45 (128 obs.) as threshold, I get a model where the data and the simulated quantiles matches nearly perfect but my data is more heavy tailed and it overestimates the rainfall for higher return levels (fig \ref{BOLT45}). If i instead pick 60 as my threshold (55 obs.), the tail and the return levels matches much better, but I get a big bump at 100 and 130 (fig \ref{BOLT60}). The distribution has parameters (68.590, 8.887, 0.423).
 		\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{BOLT45.jpg}
 		\caption{Plots with threshold = 45, BOL}
 		\label{BOLT45}
 	\end{figure}
 		\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{BOLT60.jpg}
 		\caption{Plots with threshold=60, BOL}
 		\label{BOLT60}
 	\end{figure}
 	If we look at general behaviour of the extreme values, we see opposite bahaviour as for ADA, with much more large data points after 2000 then before (fig \ref{BOLPOT}).
 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{BOLPOTplot.png}
 		\caption{Plots with threshold = 60, BOL}
 		\label{BOLPOT}
 	\end{figure}
 	\subsection{KRA}
 	The annual maxima of KRA does not really have a shape at all (fig \ref{KRAyMax}) and fits a GPD distribution with parameters (82.566, 4.789, 0.225) (fig \ref{KRAyMfit}), but with a large bump around 170.
 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{KRAyMaxPlot.png}
 		\caption{Annual maxima, KRA}
 		\label{KRAyMax}
 	\end{figure}
 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{KRAyMaxfit.jpg}
 		\caption{Plots over distribution fit for KRA annual max}
 		\label{KRAyMfit}
 	\end{figure}
 	KRA behaves very similar to BOL for the threshold 45 (204 obs.), the data has a heavier tail then the fitted distribution, but we get a nearly perfect fit for the quantiles of the simulated data and the density, but overestimates the rainfall for higher return levels (fig \ref{KRAT45}). But it did not improve very much at all by changing to 60, so instead I used 70 (66 obs.). Now both the tail behvaiour and the return levels matches much better but we see a large bump at 150 (fig \ref{KRAT70} ). The data fits to a model with parameters (82.376, 13.211, 0.564). 
 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{KRAT45.jpg}
 		\caption{Plots with threshold = 45, KRA}
 		\label{KRAT45}
 	\end{figure}
 	\begin{figure}[H]
 		\centering
 		\includegraphics[width=0.8\linewidth]{KRAT70.jpg}
 		\caption{Plots with threshold=70, KRA}
 		\label{KRAT70}
 	\end{figure}
	KRA has a similar behaviour to BOL, with more extreme days after 2000 hen before and also more spread out instead of in clusters. 
	\begin{figure}[H]
		\centering
		\includegraphics[width=0.8\linewidth]{KRAPOTplot.png}
		\caption{Plots with threshold = 70, KRA}
		\label{KRAPOT}
	\end{figure} 
	\subsection{Summary}

	\begin{table}[H]
		\centering
		\begin{tabular}{| c | c | c | c | c | c | c | c | c |}
			\hline
		 	& Obs. & Cap & \multicolumn{2}{|c|}{Gamma} & Threshold & \multicolumn{3}{|c|}{GPD} \\
		 	\hline
		 	 & & & Scale & Shape & & Location & Scale & Shape \\
		  	\hline
		  	ADA & 2037 & 100 & 17.590 & 0.585 & 50 & 60.806 & 10.714 & 0.407 \\
		 	 \hline
		 	 BOL & 2688 & 100 & 17.249 & 0.698 & 60 & 68.590 & 8.887 & 0.423 \\
		 	 \hline
		 	 KRA & 2991 & 120 & 21.774 & 0.607 & 70 & 82.376 & 13.211 & 0.564 \\
		  	\hline
		 \end{tabular}	
		\caption{Summary of parameters}
		\label{tablepara}
	\end{table}	
	I can of course not comapare some of the parameters directly since I have used different thresholds and caps, but more the shape of them. ADA and BOL have a similar Gamma distribution, with the difference that BOL has a bigger shape parameter. KRA's shape parameter is between the other two but with a larger scale parameter, possibly since I picked a larger cap. The location parameter in the GPD must be view in relation to the threshold. All of the distributions are shifted veery similarly from the threshold (11, 8, 12) and have similar shape, but differ a bit in their scale. So they do not seem to come from the same distribution.

	If I look at mean excess plots to confirm my choice of tail distribution, only the plot for KRA shows a clear linear behaviour, the other two behaves as if they should have an exponential tail (fig \ref{MeanEplot}). However, if I try to fit the upper values to an exponential distribution, it is very clear that the data is much more heavy tailed then the exponential, so I choose to stick with the GPD for all of them.
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MEplotADA.png}
			\caption{ADA, 10 points omitted}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MEplotBOL.png}
			\caption{BOL, 10 points omitted}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MEplotKRA.png}
			\caption{KRA, 9 points omitted}
		\end{subfigure}
		\caption{Mean excess plots}
		\label{MeanEplot}
	\end{figure}

	\section{Comperative assesment}
		\subsection{Data}

		\begin{table}[H]
			\centering
			\begin{tabular}{| c | c | c | c | c| c| c | c | c |}
				\hline
				Station & Long. & Lat. & Pos. Obs. & Missing values & Days (Dec, Jan, Feb) & Annual mean & Rainy day mean & Mode \\
				\hline
				AXM & -2.23 & 4.86 & 4282 & 58 & 214, 104, 154 & 1874 & 13.13 &  Semi-bi \\
				ODA & -0.98 & 5.93 & 4180 & 0 & 140, 60, 151 & 1407 & 10.10 & Bi \\
				BEK & -2.33 & 6.2 & 3851 & 366 & 80, 6, 141 & 1394 & 10.50 & \\
				KDA & -0.25 & 6.08 & 3756 & 30 & 112 , 81, 167 & 1293 & 10.33 & Bi\\
				ABE & -0.73 & 6.65 & 3415 & 700& 65, 46, 112 &1277 & 10.47 &  \\
				KSI & -1.6 & 6.71 & 3586 & 0 & 80 , 51, 136 & 1347 & 11.27 & Bi \\
				HO & 0.46 & 6.6 & 3409 & 0& 86, 52, 140 & 1276 & 11.23& Bi \\
				TDI & -1.76 & 4.88 & 3312 & 28 & 113, 37, 92 & 1079 & 9.77 & Semi-bi\\
				SUN & -2.33 & 7.33 & 3187 & 61 & 52, 23, 106 & 1191 & 11.22 & Bi\\
				WEN & -2.1 & 7.75 & 3188 & 2& 48, 24, 90 & 1249 & 11.75& Bi \\
				KRA & -0.03 & 7.81 & 2991 & 0 & 30, 20, 35 & 1366 & 13.70 & Uni\\
				BOL & -2.48 & 9.03 & 2688 & 2 &  21, 7, 29 & 1101 & 12.29 & Uni\\
				SAL & -1.06 & 5.2 & 2679 & 123 & 52, 23, 106 & 931 & 10.43 & Semi-bi\\
				NAV & -0.01 & 9.45 & 2623 & 62& 6, 6, 22 & 1024 & 11.71 & \\
				WA & -2.5 & 10.1 & 2609 & 62 & 6, 6, 22 & 1018 & 11.70 &  \\
				TLE & -0.85 & 8.5 & 2599 & 31 & 7, 6, 23 & 1017 & 11.74 & Uni\\
				AKA & 0.8 & 6.11 & 2236 & 580 & 74, 23, 56 & 848 & 10.99 & \\
				NAV1 & -1.1 & 10.9 & 2184 & 32 & 4, 2, 11 & 988 & 13.57 & Uni\\
				ACC & -0.16 & 5.6 & 2130 & 6 & 54, 29, 56 & 747 & 10.52 & Semi-bi \\
				ADA & 0.63 & 5.78 & 2037 & 0 & 37, 21, 58 & 790 & 11.63 & Semi-bi\\
				TEM & 0 & 5.61 & 1840 & 31 & 42, 23, 49 & 659 & 10.75 & Semi-bi \\
				\hline
			\end{tabular}
			\caption{All stations}
			\label{Tstations}
		\end{table}
	All of the missing values comes in clusters corresponding to months, not neccesairly back to back. The three stations with many missing values are all missing a full year. Looking at table \ref{Tstations}, I cannot see any direct pattern of the geographically position and the number of rainy days. But what is very evident is that it is a massive spread on the number of rainy days between the stations, and also the spread of number of rainy days in the dry period, even for stations with similar amount of rainy days. One way of splitting up the data could be stations with more than 3000 days of rain and stations with less, since that would split both the interval and the number of stations in half (!) i.e the number of rainy days is uniformly distributed. The annual mean sort of follows the same pattern as the number of positive observations, wheras the rainy day mean seems to follow a different pattern.

	There are many ways to split up the data to get the most accurate distribution; number of rainy days, number of rainy days in dry period, annual mean or rainy day mean. Here I have chosen to split it in half on the basis of number of days with rain, which gives the same grouping as if I split it with the annual mean as parameter. A further analysis to do, is to instead use the rainy day mean as parameter and split it. Of course the data will no longer be independet, but it was not truly independent before either.

	\begin{table}[H]
		\centering
		\begin{tabular}{|c| c c c c c c c c c c c c|} 
			\hline
			& Jan & Feb & Mar & Apr & May & Jun & Jul & Aug & Sep & Oct & Nov & Dec \\ 
			\hline\hline
			AXM & 74.8 & 71.0 & 124.2 & 132.6 & 182.2 & 209.5 & 189 & 162.8 & 110.2 & 180.1 & 112.9 & 87.9 \\
			ODA & 59.1 & 79.8 & 127 & 78.3 & 166.9 & 89 & 101.6 & 58.8 & 74.7 & 99.6 & 69.4 & 52.3 \\
			BEK & 54.3 & 86.8 & 93.6 & 95.7 & 118.4 & 104.4 & 90.2 & 85 & 117.8 & 86 & 53.9 & 154 \\
			KDA & 43 & 71.2 & 159.6 & 105.8 & 64.7 & 219.9 & 101.8 & 84.8 & 90.6 & 65.5 & 84.7 & 70.4 \\
			ABE & 65.2 & 72.6 & 91.6 & 97.8 & 79.2 & 145.6 & 77.5 & 87.5 & 121.1 & 97.5 & 51.9 & 67.6 \\
			KSI & 59.9 & 60.2 & 83.3 &  96.1 & 125.2 & 145.8 & 90 & 75 & 111.7 & 111.5 & 48.4 & 84 \\
			HO & 68.5 & 61.9 & 72.4 & 86.8 & 128.7 &  154.2 & 91.9 & 92.1 & 140.3 & 77.7 & 68.4 & 65.1 \\
			TDI & 34.5 & 77.8 & 79.8 & 90.4 & 150.5 & 124.7 & 158.4 & 82.5 & 150.7 & 152 & 73.6 & 82.9 \\
			SUN & 47.2 & 70.1 & 76.1 & 86 & 76 & 121.8 & 96 & 71.5 & 97.8 & 102.6 & 45.6 & 56 \\
			WEN & 29 & 55.6 & 143.3 & 118 & 99.4 & 118.4 & 102.2 & 76.9 & 137.5 & 94.3 & 50.2 & 32.7 \\
			KRA & 46.3 & 52.6 & 68.4 & 94.4 & 165.4 & 149 & 150.8 & 152.2 & 168.6 & 180.7 & 65.8 & 51.9 \\
			BOL & 37.1 & 79.8 & 110.3 & 81.3 & 92.6 & 101.1 & 130.1 & 122 & 139.2 & 94.7 & 74 & 45.3 \\
			SAL & 47.2 & 70.1 & 76.1 & 86 & 76 & 121.8 & 96 & 71.5 & 97.8 & 102.6 & 45.6 & 56 \\
			NAV & 36.9 & 67.9 & 40.6 & 72.5 & 126 & 113.5 & 87.7 & 132.9 & 78.3 & 68 & 18.7 & 37 \\
			WA & 36.9 & 67.9 & 40.6 & 72.5 & 126 & 113.5 & 86 & 132.9 & 101.7 & 68 & 18.7 & 37 \\
			TLE & 36.9 & 67.9 & 77.2 & 72.5 & 126 & 113.5 & 86 & 132.9 & 78.3 & 68 & 18.7 & 37 \\
			AKA & 50.8 & 43.8 & 74.4 & 116.4 & 113.2 & 90.7 & 77.2 & 78.9 & 109.4 & 77.6 & 75.6 & 70.1 \\
			NAV1 & 0.9 & 22 & 84.6 & 70.6 & 89.6 & 80.6 & 116 & 148.2 & 133.4 & 52.2 & 20.2 & 31.8 \\
			ACC & 122.5 & 71.4 & 72.3 & 124.1 & 157.9 & 123.3 & 243.9 & 57.4 & 84.6 & 150.7 & 66.8 & 123.7 \\
			ADA & 31.3 & 74.4 & 82.6 & 124.9 & 80.8 &  205.7 & 93 & 53.6& 102.1 & 100.3 & 98.4 & 65 \\ 
			TEM & 39.7 & 86.7 & 89.6 & 116.7 & 118.2 & 119.7 & 129.4 & 25.9 & 78.8 & 71.7 & 57.9 & 74.5 \\
			\hline
		\end{tabular}
		\caption{Table of yearly maximum}
		\label{tableymax}
	\end{table}	
	WA, NAV and TLE are practically identical with only a few observations differ from each other, which seems to be the dates that they do not have in common due to missing data points. So clearly these three cannot be treated as independently distributed, so I only used TLE in my analysis because it has got the fewest missing data points.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.8\linewidth]{MaxRainAll.jpg}
		\caption{Maximum value in each month, all stations}
		\label{MaxAll}
	\end{figure}
	Even if figure \ref{MaxAll} is a bit hard to follow for each station, it is possible to get a view of how the spread and the general behvaiour of the maximum values behaves. We can see that the values in January and December are pretty clusered together, except from the dashed violet which is much lower and dashed gray in January which is much higher and the solid blue and dashed gray in December which both are larger. It is also clear that the spread in April and September is much smaller with no value lying directly outside the others, and February is even closer if we ignore the dashed violet line.

	\subsection{Distribution fitting}
	To try and get as accurate distribution fits as possible, I split the data between WEN and KRA and ignored NAV and WA to not have clearly dependent data. For the higher values, I tried both 100 and 110 as splitting value for the gamma and GPD distribution, but I cannot relly see which one is the better fit for the gamma distribution(fig \ref{GammaH})

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{GammaH100.jpg}
			\caption{Data $<$ 100, gamma(17.20, 0.615)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{GammaH110.jpg}
			\caption{Data $<$ 110, gamma(17.47, 0.612)}
		\end{subfigure}
		\caption{QQ plot, gamma higher values}
		\label{GammaH}
	\end{figure}

	If we insted look at the fit of the GPD for both 100 and 110 as breaking point, it becomes quite clear that 110 is the better choice (fig \ref{GPDH})

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{HExtrFit100.jpg}
			\caption{Data $\geq$ 100, GPD(111.92, 12.43, 0.52)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{HExtrFit110.jpg}
			\caption{Data $\geq$ 110, GPD(124.60, 14.21, 0.29)}
		\end{subfigure}
		\caption{Higher values GPD fit}
		\label{GPDH}
	\end{figure}

	If I do the same analysis on the lower values, I think the break value should be either 90 or 100 (fig \ref{GammaL})

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{GammaL90.jpg}
			\caption{Data $<$ 90, gamma(18.20, 0.616)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{GammaL100.jpg}
			\caption{Data $<$ 100, gamma(18.59, 0.611)}
		\end{subfigure}
		\caption{QQ plot, gamma lower values}
		\label{GammaL}
	\end{figure}

	Now looking at the GPD fit instead, we can once again see that the higher value gives the most accurate extreme fit, and can see that more clearly if we remove the threee largest values to rescale the graphs (fig \ref{GPDL}). We can very clearly see the bump in the data around 150, which showed even when I just studied one station at a time. We can see it for the higher value stations as well, but not at all as clear. We can also see that it is worth splitting up the data since we get two different MLE distribution for the two sets if we use the same splitting value, and 100 is a suitible value for the lower data set wheras it is not very good for the higher.

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{LExtrFit90.jpg}
			\caption{Data $\geq$ 90, GPD(102.07, 12.83, 0.49)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{LExtrFit100.jpg}
			\caption{Data $\geq$ 100, GPD(114.67, 13.52, 0.35)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{LExtrFit100Min.jpg}
			\caption{Data $\geq$ 100, 3 largest values removed, GPD(114.47, 12.65, 0.22)}
		\end{subfigure}
		\caption{Lower values GPD fit}
		\label{GPDL}
	\end{figure}

	We can also look at the Mean excess plots to confirm that GPD is the correct distribution for the extreme values (fig \ref{MEplots}).
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MEH.png}
			\caption{Higher values}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MEL.png}
			\caption{Lower values}
		\end{subfigure}
		\caption{Mean excess plots}
		\label{MEplots}
	\end{figure}

	\section{Rain mode analysis}
	By plotting the monthly average of each station, we can easily see a clear uni modal, bi modal or semi-bi model behaviour for each station. We can therefor split the stations into these three categories and plot these 3 groups monthly averages instead, to get even more data to work with (fig \ref{MonAvegroups})
		\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveUni.png}
			\caption{Uni modal}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveBi.png}
			\caption{Bi modal}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{MonAveSemi.png}
			\caption{Semi bi modal}
		\end{subfigure}
		\caption{Monthly averages for each rain mode group}
		\label{MonAvegroups}
	\end{figure}

	\subsection{Uni modal stations (KRA, BOL, TLE, NAV1)}
	As many other has discovered, the rain distribution in the north only has one rainy season, as can clearly be seen in the histograms (fig \ref{MonAvegroups}). It only misses data from one november and one december and it has got proportionaly equally many positiv observations as the semi-bi modal group, but very differently distributed. It has got a much dryer Dec-Feb period then the other two groups, so we do not have enough data in these months to fit a distribution. For the other months, we have enough data to try and fit distributions. By plotting histograms for each month, one can see that all months still seems to follow a gamma distribution, alternatively a lognormal or exponential. By looking at QQ plots of the months, we can see that all months with many data points fits very well up to around 60 or 80 mm, depending on month(fig \ref{GammaFitUni}). 
	\begin{figure}[H]
		\ContinuedFloat*
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitMar.jpg}
			\caption{March,(15.633, 0.633) }
			\end{subfigure}%
			\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitApr.jpg}
			\caption{April, (19.673, 0.674)}
		\end{subfigure}
			\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitMay.jpg}
			\caption{May, (18.326, 0.705)}
		\end{subfigure}%
			\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitJun.jpg}
			\caption{June, (19.902, 0.703) }
		\end{subfigure}
			\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitJul.jpg}
			\caption{July, (19.678, 0.663)}
		\end{subfigure}%
			\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitAug.jpg}
			\caption{August, (22.687, 0.614)}
		\end{subfigure}
			\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitSep.jpg}
			\caption{September, (17.848, 0.716)}
		\end{subfigure}%
			\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{UniFitOct.jpg}
			\caption{October, (15.229, 0.726)}
		\end{subfigure}
		\caption{Gamma distribution fits, Uni modal group}
		\label{GammaFitUni}
	\end{figure}
		\begin{figure}[H]
			\ContinuedFloat
			\centering
			\begin{subfigure}{0.7\textwidth}
			\includegraphics[width=0.8\linewidth]{UniFitNov.jpg}
			\caption{November, (11.537, 0.643)}
		\end{subfigure}
		\caption{Gamma distribution fits, uni modal group}
		\end{figure}

	July is one of the clearest months that the fitted model is not suitable for the entire data set. If we instead split the data by 120 mm, the new fit works a lot better. November is a pretty poor fit as well, but we do not have enough data to make it much better. October clearly has a couple of large outliers, but fits nearly perfect else. We can most likely improve the fit of both September and March by splitting the data. 
	\begin{figure}[H]
		\centering
		\includegraphics[width=1\linewidth]{UniFitJul120.jpg}
		\caption{July, Data $<$ 120, (18.668, 0.677)}
	\end{figure}

	\subsection{Bi modal stations (ODA, KDA, KSI, HO, SUN, WEN)}

	\begin{figure}[H]
		\ContinuedFloat*
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitJan.pdf}
			\caption{January, (15.983, 0.648)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitFeb.pdf}
			\caption{February, (15.275, 0.705)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitMar.pdf}
			\caption{March,(18.644, 0.666)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitApr.pdf}
			\caption{April, (19.447, 0.737)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitMay.pdf}
			\caption{May, (18.246, 0.691)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitJun.pdf}
			\caption{June, (20.498, 0.627)}
		\end{subfigure}
		\caption{Gamma distribution fits, Uni modal group}
		\label{GammaFitBi}
	\end{figure}
	\begin{figure}[H]
		\ContinuedFloat
		\centering
			\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitJul.pdf}
			\caption{July, (19.122, 0.532)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitAug.pdf}
			\caption{August, (13.068, 0.547)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitSep.pdf}
			\caption{September, (16.732, 0.625)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{BiFitOct.pdf}
			\caption{October, (14.17, 0.73)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\includegraphics[width=1\linewidth]{BiFitNov.pdf}
			\caption{November, (11.099, 0.736)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\includegraphics[width=1\linewidth]{BiFitDec.pdf}
			\caption{December, (14.223, 0.672)}
		\end{subfigure}
		\caption{Gamma distribution fits, bi modal group}
	\end{figure}

	\subsection{Semi-bi modal stations (AXM, TDI, SAL, ACC, ADA, TEM)}

	\begin{figure}[H]
		\ContinuedFloat*
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitJan.pdf}
			\caption{January, (15.346, 0.599)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitFeb.pdf}
			\caption{February, (17.008, 0.583)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitMar.pdf}
			\caption{March,(19.727, 0.627)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitApr.pdf}
			\caption{April, (22.747, 0.588)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitMay.pdf}
			\caption{May, (24.726, 0.591)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitJun.pdf}
			\caption{June, (30.324, 0.529)}
		\end{subfigure}
		\caption{Gamma distribution fits, semi-bi modal group}
		\label{GammaFitSemi}
	\end{figure}
	\begin{figure}[H]
		\ContinuedFloat
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitJul.pdf}
			\caption{July, (21.233, 0.465)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitAug.pdf}
			\caption{August, (7.705, 0.564)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitSep.pdf}
			\caption{September, (11.671, 0.519)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{SemiFitOct.pdf}
			\caption{October, (19.981, 0.538)}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\includegraphics[width=1\linewidth]{SemiFitNov.pdf}
			\caption{November, (15.698, 0.621)}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\includegraphics[width=1\linewidth]{SemiFitDec.pdf}
			\caption{December, (16.805, 0.575)}
		\end{subfigure}
		\caption{Gamma distribution fits, semi-bi modal group}
	\end{figure}

	\section{Justification of splitting and model selection}
	To find the most suitable distribution for the data, I started with the threshold 80 to eliminate at least all extreme values, but still keep large values to not change the structure of the data to much. I then fit this smaller data set to; gamma, log-normal and exponential distribution, since all of them where suggested models in other rain prediction papers. I used the AIC, by taking the model with lowest value, to determine the most suitable model, which was the gamma distribution. I then fit the data both using MLE and MGE method and concluded that MLE was the best method, once again with use of AIC.

	To determine if the model became better if I only used data up to a certain value, and what value that should be, I instead used Q-Q plots. I fit the gamma distribution to data sets smaller than 40, 50, 60, 70, 80, 90 and 100 mm per day and the full data set. It is very clear that we get a much better model fit if we exclude the most extreme values. The same gamma distribution is fitted if I use 80,90 or 100 as the splitting value, probably since I add very few values in each step.

	\section{Wilcoxon}
	By splitting up the data in different ways; in year groups, by annual rainfall or average daliy rainfall, we can use Wilcoxon sign rank test to see if these groups come from the same distribution of if they are different. If we split up the data in decades, and get the average rainfall in each month , we can treat these two data sets as paired. The test is not significant if we run it with consecutive decades (0.30, 0.42) but it is significant on a 10\% level between the first decade and the last (0.09). This gives a very small proof that the rain pattern might have changed over the 30 years, but you would need more years to prove such a statement, since you have easily have fluctations in this period.

	If we instead calculate the average monthly rainfall between the stations with highest rainfall and lowest(values $<$ 80), we instead get a strongly significant difference with a p-value = 0.006836. So clearly there is a stronger variability between the stations then between the decades.

	\section{Literature review}
	In general, I have had a hard time finding any work on rainfall distributions in Ghana, but it has been done in other tropical regions and similar work has been done in temperature. Work in Ghana has mainly been focused on the variability in annual or seasonal rainfall.

	Ghana is a country in the south of Africas horn, by the coast and shares boarder with Burkina Faso, C\^{o}te d'Ivoire and Togo. It has five distinct geographical areas; low plains in the south, the Volta Basin in the centre with the artificial lake 'Lake Volta', the Akwapim-Togo ranges to the east of the Volta Basin with many heights and folded strata, the Ashanti Uplands to the west and high plains in the north\cite{fao}. The temperature is peaking around February-March and at its lowest around August. Ghana has three distinct rainfall behvaiours. The northern part experiences an unimodal season with the rainy season between April and September, wheras the rest of the country has a bimodal seanson, first one in April to July and the second September to November. The difference is that some parts of the country has two modes of the same amplitude wheras the other has peaks of different aplitudes. But they all have in common a slowly increasing peak but a rapid decrease in October.\cite{RainVarGhana} The different rain patterns depend on a few wind and preassure phenomenons. It is strongly affected by the  position of the \textbf{Inter-tropcial convergence zone}, which goes between the nothern and southern tropics every year. The prevailing winds north of the ITCZ is called the Harmmatan and brings hot and dusty air from the Sahara desert between Dacember and March, which gives rise to the very dry season. The prevailing wind south of ITCZ is southwesterly and instead brings humid air from the Atlantic ocean. As ITCZ moves from the nothern position to the souther and back, the opposing prevailing winds gives rise to the West African monsoon which shows as the two rainy seasons. The rainy season in the north corresponds to when the ITCZ are at its most nothern position\cite{RainVarGhana}.

	To get a better idea of how the proportion of wet and rainy days are distributed several ideas are proposed. In \cite{MalaysiaBin} they use the binomial distribution to model dry and wet days and combine it with a continuose distribution for the wet days. The continuouse distributions they look at are; gamma, lognormal, exponential and weibull, which seems to be the standard distributions to test. They determined which was the most suitable model by looking at the AIC and picked the model with the lowest value. They concluded that the most stations fitted best to a mixed lognormal but some to the gamma, and its was strongly connected with the surrouding topology.

	\cite{MixedDistr} they instead fitted mixted distributions, so two distributions of the same type but with different parameters, and then a weight parameter for the two distributions. This was also done on Malyasian daily rainfall, and they used MLE to fit the distribution. To then identify the most suitable model, they used 7 different goodness-of-fit tests; median absolute difference, Kolmogorov-Smirov, Cramer-von-Mises, Anderson-Darling, New Kolmogorov-smirnov, New Cramer-von-Mises, New Anderson-darling. Since they used so many test, they determined which was the best model by taking the model that performed best in most tests, and did not do any analysis on which tests where the most suitable. Most stations fitted to a mixed weibull and a few to a mixed gamma, once again stronlgy influenced by their geography and topology. Finally, they looked at what model fitted best for each monsoon season and their transitional periods.

	In \cite{Markov} they instead use a Markov chain to look at the probability of rain given a certain number of dry days, and let that probability change with months because of seasonality. This gives a much more descriptive representation of the distribution of rain. They also decided to classify rain less the 2.5 mm as trace instead and made seperate probabilities for that.

	A completly different approach is used by \cite{Bayesian}, where they instead of trying to fit a model to the data points, treat the data points as realisations of a multivariate normal distribution. The data they are working with is measured in mm and does not classify any rain as "traces". This method tries to predict rainfall in both space and time, by letting the observation be a parameter dependent on both. They base their model on a truncated normal model
	\begin{equation}
	z =
	\begin{cases}
	w^\beta,& \text{if} \ w > 0 \\
	0, & \text{if} \ w  \leq 0 
	\end{cases}
	\end{equation}
	where z is the observed rain fall at a specific station and time and $w$ is distributed normally with a known mean and variance. By using a Bayesian method, they incorporate the inceartanty of the parameters into the posterior distribution.

	\section{West African metorology}
	The monsoon differs over Africa due to its geographical difference. In west Africa, the northern part consists of land and the southern part of ocean wheras east Africa only consists of land, eventhough the northern part is wider then the southern. In the northern hemisphere summer a thermal low pressure builds up over the continent around 20\textdegree N and the ITCZ moves northward to 15\textdegree N. At the same time, heat troughs develop over North Atlantic ocean, the anticyclone system \textbf{St Helena} builds up ove the South Atlantic ocean and cold water is flowing northward along the south-west African coast. Beacause of all this a pressure gradient is formed between South Atlantic and north Africa which makes the south-easterly flows to recurve and become the south-westerly monsoon flow over west Africa.(McGregor:1998). This flow brings cold and moist oceanic air which is the reason for rain, so the area with maximum rainfall is where we have thickest air masses, i.e closest to the coast. North of the ITCSZ conditions are genrally cloudless and dry and south of the zone we find the most cloud cover and maximum rain. Dry years are coused by: retardation of the northward movement of the south-west monsoon, a southward displacemnt of the intertropical discontinuity,the near-equatorial trough and the xone of maximum surface pressure and anoalously cold water to the nort-west of the line linking SW West Africa nd NE Brasil. 
	During the northern hemisphere winter, West Africa is instead under the influence of  north-easterly trade winds which bring dry and stable air masses, often containing dust particles. These winds are called \textbf{Harmattan}. At this time, the ITCZ is located to the south of the southern west-African coast, hence why there is no rain over west Africa this time of year.

	It is not clear wheather ENSO has an impact on the African monsoon or not, but it seems that very intense warm phases of it can reduce the monsoon precipitation, as recorded during the -83 ENSO. The \textbf{west African mid-Troposhperic jet} is a thermal wind that occurce due to the temperature gradient between the warm Sahara and the cold Gulf of Guinea. We can find maximum rainfall on the equatorward side of the jet. The \textbf{ITCZ} is composed of two zones, one with low pressure, maximum surface temperature and wind confluence and another zone with maximum cloudiness, therefore also maximum rainfall and wind converge. These two zones can be as far aprat as 1000 km. 
	\textbf{Squall lines} is a linear system of many thunderstorms or clouds that behave as one and therefore can live for 13-15 h instead of just a couple of hourse. They can be hundreds of km long and 30 km wide and are characterised by their explosive growth, rapid propagation and their convex leading ege(?). They peak in the afternoon and usually form west of mountains which implies that surface heating and orographic  effects creates them. In west Africa, the African easterly jet seems to have a big influence on the initiation of them. Moisture in the atmosphere is a key thing for creating squall lines, by having moist lower levels and dry mid levels a postive feedback is created over the zone and therfore further developes the zone. There are more squall lines during the beginning and end of the monsoon because the dry mid level is absent during the middle of the monsoon.

	Clouds are generally on a higher altitude in the tropics compared to the northern hemisphere. All rainfall is a result of upward movements of moist air. For this to happen, the atmosphere must be in a state of conditional, potential or convective instability. Three types of rainfall can then occure: convectional, cyclonic or orographic. The tropics only experiences convectional rainfall which is characterised by short and intense rainfall. Months with more then 50 mm of rain are sufficiently rainy for crops to grow without irrigation.

	\section{Dictionary}
	\begin{itemize}
		\item \textbf{Convergence}: Difference in wind speed forces air to "pile up", which creates a vertical movemnet upwards or downwards depending if the convergence is on the surface or up in the atmosphere.
		\item \textbf{Confluence}: Difference in wind speed gets air together but as wind enters the confluence zone, it speeds up, hence does not converge.
		\item \textbf{Baroclinic}: A atmosphere for which the density depends on both temperature and pressure.
		\item \textbf{Barotropic}: A atmosphere for which the density only depends on pressure.
		\item \textbf{Wind shear}: Change in wind speed and/or direction over a short distance in the atmosphere. It can be both horisontaly or verticaly and usually observed close to weather fronts or thermal winds.
		\item \textbf{Trough (dal)}: Extended region of relatively low atmospheric pressure.
		\item \textbf{Latent heat}: Energy transfer to the atmosphere due to evaporation.
		\item \textbf{Relative humidity}: ratio of amount of water present it the atmosphere realtive to the amount that could be present at that given temperature.
		\item \textbf{Dew formation(dagg)}: Condensation on cool surfaces.
		\item \textbf{Anabatic}: Upward wind motion.
		\item \textbf{Adiabatic}: Any change in internal energy only depends on work such as compression or expansion. Opposite is \textbf{diabatic} which is a change in energy between a system and its surrounding due to a temperature gradient.
		\item \textbf{Sensitivity}: An evaluation of how much each input contributes to the model output uncertainty. This can be done in numerous of ways, linear regression being one or running the model and changing one variable at a time.
	\end{itemize}

	\section{Comparing data to CMIP5}
	\subsection{Rain over a threshold}
	Studying figure \ref{PRCPTOT}, \ref{R1mm}, \ref{R10mm} and \ref{R20mm} we can see a very clear pattern in the difference between our data and the GCM (Global Climate Models). The GCMs heavily over estimates the number of rainy days which also leads to a vast over estimation of the annual total rainfall (figure \ref{PRCPTOT}, \ref{R1mm}). But the span among the GCMs is massive, ranging from 600 mm up to 2800 mm per year, wheras the range when splitting the data into modes is only between 500 and 1500 mm per year. So it appears that at least a few of the models simulates in the correct range, but looking at the CMIP5 mean it is evient that the majority of the GCMs simulates a much to high annual rainfall. However, the behaviour of the data curve and the CMIP5 mean curve is similar, which could mean that the GCMs can correctly simulate the changes between years even if they cannot correctly  simulate number of rainy days.

	Looking at figure \ref{R1mm}, we can see that all models simulate more rainy days than our data, which leads to the mean curve to be about 100 days per year shifted compare to the data curve. But the range between the different GCMs is once a gain large, ranging from 100-300 days. Looking at the CMIP5 mean curve, we can see that the models are not as good at simulating the variation in rainy days between years as they are at simulating changes in rain amount.

	When looking at heavier rainfall, the second known issue becomes clear. The GCMs can simulate the number of days with $\geq$ 10 mm fairly well, the curve is slightly higher then our data but not completly out of range, wheras for $\geq$ 20 mm, they simulate too few days. This is a well know problem, the over simulation of rainy days and the lack of skill to simulate days with very heavy rainfall. Another interesting thing to notice is that CMIP5 mean curve seems to have a uppward pointing trend for both $\geq$ 10mm and $\geq$ 20 mm which is not clear in our data. This is a big issue if we want to use these models to predict future behaviour or precipitation.

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{PRCPTOTall.pdf}
			\caption{All stations}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{PRCPTOTmodes.pdf}
			\caption{Modes}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{PRCPTOTGCMall.pdf}
		\caption{CMIP5 all}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{PRCPTOTGCMmean.pdf}
		\caption{CMIP5 mean}
	\end{subfigure}
		\caption{Total annual precipitation ($\geq$1 mm)}
		\label{PRCPTOT}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R1mmall.pdf}
			\caption{All stations}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R1mmodes.pdf}
			\caption{Modes}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R1mmGCMall.pdf}
			\caption{CMIP5 all}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R1mmGCMmean.pdf}
			\caption{CMIP5 mean}
		\end{subfigure}
		\caption{Number of rainy days ($\geq$1 mm)}
		\label{R1mm}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R10mmall.pdf}
			\caption{All stations}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R10mmodes.pdf}
			\caption{Modes}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R10GCMall.pdf}
			\caption{CMIP5 all}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R10GCMmean.pdf}
			\caption{CMIP5 mean}
		\end{subfigure}
		\caption{Number of rainy days ($\geq$10 mm)}
		\label{R10mm}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R20mmall.pdf}
			\caption{All stations}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R20mmodes.pdf}
			\caption{Modes}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R20GCMall.pdf}
			\caption{CMIP5 all}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R20GCMmean.pdf}
			\caption{CMIP5 mean}
		\end{subfigure}
		\caption{Number of rainy days ($\geq$20 mm)}
		\label{R20mm}
	\end{figure}

	\subsection{Rain quantiles}
	Since CMIP5 uses earlier years then what we have data from as reference period, it is not possible to compare the numbers with each other, but more the behaviour od the curve. 84-93 is picked as a reference period instead of 83-92 to avoid the clearly lower values in -83. In figure \ref{R95} both curves seems to exhibit a very similar behaviour, which is a steady increase in the total rainfall on days with heavy rainfall. For days with extreme rainfall (figure \ref{R99}), the simulated mean is very close to the data mean, but the spread among the models are still very large. The CMIP5 mean is again showing a steady inscrease which is not clearly visible in the data plot. So similar differences can been seen both when looking at the highest percentiles and very heavy rainfall in mm. 
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R95all.pdf}
			\caption{All stations}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R95GCMall.pdf}
			\caption{CMIP5 all}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R95GCMmean.pdf}
			\caption{CMIP5 mean}
		\end{subfigure}
		\caption{Total rain amount in days above 95\% threshold for reference period}
		\label{R95}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R99all.pdf}
			\caption{All stations}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R99GCMall.pdf}
			\caption{CMIP5 all}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{R99GCMmean.pdf}
			\caption{CMIP5 mean}
		\end{subfigure}
		\caption{Total rain amount in days above 99\% threshold for reference period}
		\label{R99}
	\end{figure}

	\section{Trends in timeseries, Lowess}
	Studying figure \ref{lowess}, it is quite evident that the annual precipitation, very heavy rainfall and the rain amount on R95 (days with more rain then the 95 percentail in the reference period), is increasing over the 30 year period. Number of rainy days is showing a small increase aswell. Number of days with $\geq$ 10 mm is not showing a consistent pattern since it is a small decrease for the first half ot period, to then increase back to the same level as 1983. R99 is showing the opposite pattern, an increase in the amount for the first half of the period to then decrease bac to the 1983 level. 
	This could suggest that there is an increase in the number of rainy days and an increase in days with very hevay rain but a slight decrese in the most extreme rainfalls. 
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{lowessAp.pdf}
			\caption{Annual precipitation}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{lowessRd.pdf}
		\caption{Number of rainy days}
	\end{subfigure}	
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{lowessR10.pdf}
		\caption{Days $\geq$ 10mm}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{lowessR20.pdf}
		\caption{Days $\geq$ 20mm}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=1\linewidth]{lowessR95.pdf}
		\caption{Days $\geq$ 95\%}
	\end{subfigure}%
	\begin{subfigure}{0.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{lowessR99.pdf}
		\caption{Days $\geq$ 99\%}
	\end{subfigure}
	\caption{Smooting using lowess}
	\label{lowess}
	\end{figure}

	If we leave the changes in extreme indicies and instead look at some extreme values for each year, figure \ref{extRainRatio} shows that the ratio of very heavy rainfall ($\geq$ 95\% quantile) is more or less constant with large flucations around -98. The extreme rainfall ratio is instead showing a downward trend. Since figure \ref{lowess} (a) show that annual precipitation is increasing over time, this indicated that the extreme rainfall is either constant or not increasing in the same tempo as the total annual rainfall. Autocorrelation plot is insignificant for both time series. 
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{extRainRatio95.pdf}
			\caption{95\% quantile}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=\linewidth]{extRainRatio99.pdf}
			\caption{99\% quantile}
		\end{subfigure}	
		\caption{Ratio extreme rainfall to total annual rainfall}
		\label{extRainRatio}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsPRCPTOTu.pdf}
			\caption{Uni}
		\end{subfigure}%
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsPRCPTOTb.pdf}
			\caption{Bi}
		\end{subfigure}%
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsPRCPTOTs.pdf}
			\caption{Semi}
		\end{subfigure}
		\caption{Total annual precipitation}
		\label{modesPRCPTOT}
	\end{figure}

		\begin{figure}[H]
		\centering
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR1u.pdf}
			\caption{Uni}
		\end{subfigure}%
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR1b.pdf}
			\caption{Bi}
		\end{subfigure}%
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR1s.pdf}
			\caption{Semi}
		\end{subfigure}
		\caption{Total number of rainy days}
		\label{modesR1}
	\end{figure}

		\begin{figure}[H]
		\centering
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR10u.pdf}
			\caption{Uni}
		\end{subfigure}%
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR10b.pdf}
			\caption{Bi}
		\end{subfigure}%
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR10s.pdf}
			\caption{Semi}
		\end{subfigure}
		\caption{Total number of days $\geq$ 10 mm}
		\label{modesR10}
	\end{figure}

		\begin{figure}[H]
		\centering
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR20u.pdf}
			\caption{Uni}
		\end{subfigure}%
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR20b.pdf}
			\caption{Bi}
		\end{subfigure}%
		\begin{subfigure}{0.3\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsR20s.pdf}
			\caption{Semi}
		\end{subfigure}
		\caption{Total number of days $\geq$ 20 mm}
		\label{modesR20}
	\end{figure}

	\section{Autocorrelation}
	Fot the threshold based inidicies there is a weak neagative autocorrelation between consecutive years, with R10 being close to significant. For the quantile based indicies there is instead a positive autocorrelation with a 6 year lag (el ni\~{n}o?).

	\begin{thebibliography}{1}
		\bibitem{Embrechts}
		Embrechts, Paul, Kl\"{u}pperberg, Claudia, Mikosh, Thomas(1997)
		\textit{Modelling extreme events for Insurance and finance},
		New york: Springer-Verlag Berlin Heidelberg
		\bibitem{fao}
		Food and agriculture organisation of the United nations(2005) \textit{Irrigations in Africa in figures - AQUASTAT survey 2005}, Available at: http://www.fao.org/nr/water/aquastat/countries\_regions/GHA/GHA-CP\_eng.pdf [Accessed 23 Feb. 2018]
		\bibitem{PennState}
		Available at:
		https://onlinecourses.science.psu.edu/stat414/node/319 [Accessed 28 Feb. 2018]
		\bibitem{RainVarGhana}
		Nkrumah, F., et al. (2014)\textit{ Rainfall Variability over Ghana: Model versus Rain Gauge Observation}. International Journal of Geosciences, 5, 673-683. http://dx.doi.org/10.4236/ijg.2014.57060
		\bibitem{MalaysiaBin}
		Open Journal of Modern Hydrology, 2011, 1, 11-22
		doi:10.4236/ojmh.2011.12002 Published Online October 2011 (http://www.SciRP.org/journal/ojmh)
		\bibitem{Markov}
		Stern, R., \& Coe, R. (1984). \textit{A Model Fitting Analysis of Daily Rainfall Data}. Journal of the Royal Statistical Society. Series A (General), 147(1), 1-34. doi:10.2307/2981736
		\bibitem{MixedDistr}
		Jamaludin, Shariffah \& Aziz Jemain, Abdul. (2007).\textit{ Fitting daily rainfall amount in Peninsular Malaysia using several types of Exponential distributions}. Journal of Applied Sciences Research. 3. 
		\bibitem{Bayesian}
		Sanso, B., \& Guenni, L. (1999). \textit{Venezuelan Rainfall Data Analysed by Using a Bayesian Space-Time Model}. Journal of the Royal Statistical Society. Series C (Applied Statistics), 48(3), 345-362. Retrieved from http://www.jstor.org/stable/2680829
	\end{thebibliography}

\begin{thebibliography}{1}
	\bibitem{Embrechts}
	Embrechts, Paul, Kl\"{u}pperberg, Claudia, Mikosh, Thomas(1997)
	\textit{Modelling extreme events for Insurance and finance},
	New york: Springer-Verlag Berlin Heidelberg
	\bibitem{fao}
	Food and agriculture organisation of the United nations(2005) \textit{Irrigations in Africa in figures - AQUASTAT survey 2005}, Available at: http://www.fao.org/nr/water/aquastat/countries\_regions/GHA/GHA-CP\_eng.pdf [Accessed 23 Feb. 2018]
	\bibitem{PennState}
	Available at:
	https://onlinecourses.science.psu.edu/stat414/node/319 [Accessed 28 Feb. 2018]
	\bibitem{RainVarGhana}
	Nkrumah, F., et al. (2014)\textit{ Rainfall Variability over Ghana: Model versus Rain Gauge Observation}. International Journal of Geosciences, 5, 673-683. http://dx.doi.org/10.4236/ijg.2014.57060
	\bibitem{MalaysiaBin}
	Open Journal of Modern Hydrology, 2011, 1, 11-22
	doi:10.4236/ojmh.2011.12002 Published Online October 2011 (http://www.SciRP.org/journal/ojmh)
	\bibitem{Markov}
	Stern, R., \& Coe, R. (1984). \textit{A Model Fitting Analysis of Daily Rainfall Data}. Journal of the Royal Statistical Society. Series A (General), 147(1), 1-34. doi:10.2307/2981736
	\bibitem{MixedDistr}
	Jamaludin, Shariffah \& Aziz Jemain, Abdul. (2007).\textit{ Fitting daily rainfall amount in Peninsular Malaysia using several types of Exponential distributions}. Journal of Applied Sciences Research. 3. 
	\bibitem{Bayesian}
	Sanso, B., \& Guenni, L. (1999). \textit{Venezuelan Rainfall Data Analysed by Using a Bayesian Space-Time Model}. Journal of the Royal Statistical Society. Series C (Applied Statistics), 48(3), 345-362. Retrieved from http://www.jstor.org/stable/2680829
	\bibitem{MTM}
	Deidda, R.: \textit{A multiple threshold method for fitting the generalized Pareto distribution to rainfall time series}, Hydrol. Earth Syst. Sci., 14, 2559-2575, https://doi.org/10.5194/hess-14-2559-2010, 2010.
	\bibitem{entirerange}
	Li, C., V. P. Singh, and A. K. Mishra (2012), \textit{Simulation of the entire range of daily precipitation using a hybrid probability distribution}, Water Resour. Res., 48, W03521, doi:10.1029/2011WR011446
	\bibitem{large p-values}
	http://blog.minitab.com/blog/statistics-and-quality-data-analysis/large-samples-too-much-of-a-good-thing [Accessed 14 March 2018]
	\bibitem{KS-test}
	https://onlinecourses.science.psu.edu/stat414/node/322

\end{thebibliography}

\end{document}

### 2018-05-18 BWmonths.R

@@ -1,0 +1,82 @@
+year <- seq(1983, 2012)
+month <- c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")
+
+
+semi.dec.c <- numeric(30)
+for (i in 1:30) {
+  semi.dec.c[i] <- sum(subset(Semi.dec$Data, year==year[i]))/6
+}
+semi.nov.c[1] <-sum(subset(Semi.nov$Data, year==year[1]))/5
+
+bxs <-boxplot(Uni.jan.c, Uni.feb.c, Uni.mar.c, Uni.apr.c, Uni.may.c, Uni.jun.c, Uni.jul.c, Uni.aug.c, Uni.sep.c, Uni.oct.c, Uni.nov.c, Uni.dec.c, main = "Uni modal, normal", names = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"))
+
+Semi.jan.c <- numeric(30)
+ for (i in 1:30) {
+    Semi.jan.c[i] <- sum(subset(Semi.jan$Data, year==year[i]))/6
+  }
+ Semi.feb.c <- numeric(30)
+ for (i in 1:30) {
+    Semi.feb.c[i] <- sum(subset(Semi.feb$Data, year==year[i]))/6
+   }
+ Semi.mar.c <- numeric(30)
+for (i in 1:30) {
+     Semi.mar.c[i] <- sum(subset(Semi.mar$Data, year==year[i]))/6
+   }
+ Semi.apr.c <- numeric(30)
+ for (i in 1:30) {
+     Semi.apr.c[i] <- sum(subset(Semi.apr$Data, year==year[i]))/6
+   }
+Semi.may.c <- numeric(30)
+ for (i in 1:30) {
+     Semi.may.c[i] <- sum(subset(Semi.may$Data, year==year[i]))/6
+   }
+ Semi.jun.c <- numeric(30)
+ for (i in 1:30) {
+     Semi.jun.c[i] <- sum(subset(Semi.jun$Data, year==year[i]))/6
+   }
+ Semi.jul.c <- numeric(30)
+ for (i in 1:30) {
+     Semi.jul.c[i] <- sum(subset(Semi.jul$Data, year==year[i]))/6
+   }
+Semi.aug.c <- numeric(30)
+ for (i in 1:30) {
[…]

### 2018-05-18 Basic functions.R

@@ -1,0 +1,73 @@
+#Read data
+data <- read.table('Ghana_daily.txt', header=TRUE)
+dates <- as.Date(as.character(data$Date), format='%Y%m%d')
+
+data.pos <- subset(data, Data>0)
+data.norm <- subset(data, Data >= 1)
+#exclude dependent and many missing data points
+data.norm <- subset(data.norm, Station=='23001AXM' | Station=='21088ODA' | Station=='22050KDA' | Station=='17009KSI' |Station=='07017HO' | Station=='23003TDI' | Station=='01032SUN'| Station=='01018WEN'|Station=='07008KRA' | Station=='07000BOL' | Station=='23022SAL' | Station=='07006TLE' | Station=='04003NAV1' | Station=='23016ACC' | Station=='23002ADA' | Station=='23024TEM')
+
+#Pick out date parts
+dfdate <- data.frame(date=dates)
+year=as.numeric (format(dates,"%Y"))
+month=as.numeric (format(dates,"%m"))
+day=as.numeric (format(dates,"%d"))
+
+dataNew <- cbind(dfdate,year,month,day, data)
+dataNew.norm <- subset(dataNew, Data >= 1)
+#exclude dependent and many missing data points
+dataNew.norm <- subset(dataNew.norm, Station=='23001AXM' | Station=='21088ODA' | Station=='22050KDA' | Station=='17009KSI' |Station=='07017HO' | Station=='23003TDI' | Station=='01032SUN'| Station=='01018WEN'|Station=='07008KRA' | Station=='07000BOL' | Station=='23022SAL' | Station=='07006TLE' | Station=='04003NAV1' | Station=='23016ACC' | Station=='23002ADA' | Station=='23024TEM')
+
+
+#Extract all values 
+unique(data$Station)
+unique(data$Long)
+
+#Transform dates
+d.TLE <- as.Date(as.character(TLE$Date), format='%Y%m%d')
+
+plot(KRA$Data~d.KRA)
+hist(data.norm$Data, breaks = 40, freq = FALSE, main = "Hist all pos. data")
+boxplot(data.pos$Data, main="Rainy days")
+
+#Subsetting years and months
+y2000 <-subset(data, substr(data$Date,1, 4)==2000)
+d.y2000 <- as.Date(as.character(y2000$Date), format='%Y%m%d')
[…]

### 2018-05-18 BulkAnalysis.R

@@ -1,0 +1,66 @@
+#Fit different distributions
+fit.gamma.mle <- fitdist(data.norm$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+fit.gamma.mge <- fitdist(data.norm$Data, "gamma",  method = "mge", gof = "CvM")
+#gives nearly identical results, mle slightly better with a little bit lower AIC
+
+summary(fit.gamma.mle)
+summary(fit.gamma.mge)
+plot(fit.gamma.mle)
+plot(fit.gamma.mge)
+
+fit.lognorm <- fitdist(data.norm$Data, "lnorm", method = "mle")
+fit.exp <- fitdist(data.norm$Data, dexp, method = "mle")
+
+summary(fit.lognorm)
+summary(fit.exp)
+
+#Different thresholds
+norm40 <- subset(data.pos, Data < 40)
+fit.gamma40 <- fitdist(norm40$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+norm50 <- subset(data.pos, Data < 50)
+fit.gamma50 <- fitdist(norm50$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+norm60 <- subset(data.pos, Data < 60)
+fit.gamma60 <- fitdist(norm60$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+norm70 <- subset(data.pos, Data < 70)
+fit.gamma70 <- fitdist(norm70$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+norm80 <- subset(data.pos, Data < 80)
+fit.gamma80 <- fitdist(norm80$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+norm90 <- subset(data.pos, Data < 90)
+fit.gamma90 <- fitdist(norm90$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+norm100 <- subset(data.pos, Data < 100)
+fit.gamma100 <- fitdist(norm100$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+norm150 <- subset(data.pos, Data < 150)
[…]

### 2018-05-18 DecadalAnalysis.R

@@ -1,0 +1,145 @@
+#Split into decades
+y8392 <- subset(dataNew.pos, year==1983 |year==1984|year==1985|year==1986|year==1987|year==1988|year==1989|year==1990|year==1991|year==1992)
+y9302 <- subset(dataNew.pos, year==1993|year==1994|year==1995|year==1996|year==1997|year==1998|year==1999|year==2000|year==2001|year==2002)
+y0312 <- subset(dataNew.pos, year==2003|year==2004|year==2005|year==2006|year==2007|year==2008|year==2009|year==2010|year==2011|year==2012)
+
+#Pick out norm values
+y8392.norm <- subset(y8392, Data <80)
+y9302.norm <- subset(y9302, Data < 80)
+y0312.norm <- subset(y0312, Data <80)
+
+#Months y8392
+y8392.norm.jan <- subset(y8392.norm, substr(y8392.norm$Date, 5, 6)=='01')
+d.y8392.normjan <- as.Date(as.character(y8392.norm.jan$Date), format='%Y%m%d')
+y8392.norm.feb <- subset(y8392.norm, substr(y8392.norm$Date, 5, 6)=='02')
+d.y8392.normfeb <- as.Date(as.character(y8392.norm.feb$Date), format='%Y%m%d')
+y8392.norm.mar <- subset(y8392.norm, substr(y8392.norm$Date, 5, 6)=='03')
+d.y8392.normmar <- as.Date(as.character(y8392.norm.mar$Date), format='%Y%m%d')
+y8392.norm.apr <- subset(y8392.norm, substr(y8392.norm$Date, 5, 6)=='04')
+d.y8392.normapr <- as.Date(as.character(y8392.norm.apr$Date), format='%Y%m%d')
+y8392.norm.may <- subset(y8392.norm, substr(y8392.norm$Date, 5, 6)=='05')
+d.y8392.normmay <- as.Date(as.character(y8392.norm.may$Date), format='%Y%m%d')
+y8392.norm.jun <- subset(y8392.norm, substr(y8392.norm$Date, 5, 6)=='06')
+d.y8392.normjun <- as.Date(as.character(y8392.norm.jun$Date), format='%Y%m%d')
+y8392.norm.jul <- subset(y8392.norm, substr(y8392.norm$Date, 5, 6)=='07')
[…]

### 2018-05-18 Histograms.R

@@ -1,0 +1,34 @@
+HistA <- hist(data.norm$Data, freq = FALSE)
+HistH <- hist(higher.norm$Data, freq = FALSE)
+HistL <- hist(lower.norm$Data, freq = FALSE)
+
+#plot(0,0, type='n', xlim=c(0,80), ylim = c(0,0.20), xlab = "rain mm", ylab = "density", main = "Density plot")
+plot(HistA, col="red", density = 10)
+plot(HistH, col="green", density = 10, add=TRUE)
+plot(HistL, col="blue", density = 10, add=TRUE)
+
+## calculate the range of the graph
+xlim <- range(HistA$breaks,HistH$breaks, HistL$breaks)
+ylim <- range(0,HistA$density,
+              HistH$density, HistL$density)
+## plot the first graph
+plot(HistH,xlim = xlim, ylim = ylim,
+     col = rgb(0.6, 0, 0, alpha = 0.3),xlab = 'Rain mm',
+     freq = FALSE, ## relative, not absolute frequency
+     main = 'Histogram of rainfall')
+## plot the second graph on top of this
+opar <- par(new = FALSE)
+plot(HistA,xlim = xlim, ylim = ylim,
+     xaxt = 'n', yaxt = 'n', ## don't add axes
+     col = rgb(0, 0.4, 0, alpha = 0.3), add = TRUE,
+     freq = FALSE) ## relative, not absolute frequency
+## add a legend in the corner
+plot(HistL,xlim = xlim, ylim = ylim,
+     xaxt = 'n', yaxt = 'n', ## don't add axes
+     col = rgb(0, 0.2, 0.3,alpha =  0.3), add = TRUE,
+     freq = FALSE) ## relative, not absolute frequency
+## add a legend in the corner
+legend('topright',c('High values','All values', 'Low values'),
+       fill = c('red', 'blue', 'green'), bty = 'n',
+       border = NA)
+par(opar)

### 2018-05-18 Means.R

@@ -1,0 +1,43 @@
+ABE.mean <- sum(ABE$Data)/length(ABE$Data)
+ACC.mean <- sum(ACC$Data)/length(ACC$Data)
+AKA.mean <- sum(AKA$Data)/length(AKA$Data)
+ADA.mean <- sum(ADA$Data)/length(ADA$Data)
+AXM.mean <- sum(AXM$Data)/length(AXM$Data)
+BEK.mean <- sum(BEK$Data)/length(BEK$Data)
+BOL.mean <- sum(BOL$Data)/length(BOL$Data)
+HO.mean <- sum(HO$Data)/length(HO$Data)
+KDA.mean <- sum(KDA$Data)/length(KDA$Data)
+KRA.mean <- sum(KRA$Data)/length(KRA$Data)
+KSI.mean <- sum(KSI$Data)/length(KSI$Data)
+NAV.mean <- sum(NAV$Data)/length(NAV$Data)
+NAV1.mean <- sum(NAV1$Data)/length(NAV1$Data)
+ODA.mean <- sum(ODA$Data)/length(ODA$Data)
+SAL.mean <- sum(SAL$Data)/length(SAL$Data)
+SUN.mean <- sum(SUN$Data)/length(SUN$Data)
+TDI.mean <- sum(TDI$Data)/length(TDI$Data)
+TEM.mean <- sum(TEM$Data)/length(TEM$Data)
+TLE.mean <- sum(TLE$Data)/length(TLE$Data)
+WA.mean <- sum(WA$Data)/length(WA$Data)
+WEN.mean <- sum(WEN$Data)/length(WEN$Data)
+
+ABE.Amean <- sum(ABE$Data)/28
+ACC.Amean <- sum(ACC$Data)/30
+AKA.Amean <- sum(AKA$Data)/29
+ADA.Amean <- sum(ADA$Data)/30
+AXM.Amean <- sum(AXM$Data)/30
+BEK.Amean <- sum(BEK$Data)/29
+BOL.Amean <- sum(BOL$Data)/30
+HO.Amean <- sum(HO$Data)/30
+KDA.Amean <- sum(KDA$Data)/30
+KRA.Amean <- sum(KRA$Data)/30
+KSI.Amean <- sum(KSI$Data)/30
+NAV.Amean <- sum(NAV$Data)/30
+NAV1.Amean <- sum(NAV1$Data)/30
[…]

### 2018-05-18 compPlots.R

@@ -1,0 +1,61 @@
+m <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
+KRA.max <- c(46.3, 52.6, 68.4, 94.4, 165.4, 149, 150.8, 152.2, 168.6, 180.7, 65.8, 51.9)
+ADA.max <- c(31.3, 74.4, 82.6, 124.9, 80.8,  205.7, 93, 53.6, 102.1, 100.3, 98.4, 65)
+BOL.max <- c(37.1, 79.8, 110.3, 81.3, 92.6, 101.1, 130.1, 122, 139.2, 94.7, 74, 45.3)
+months <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)
+
+plot(months, ADA.max, type = "o",col = "red", xlab = "Month", ylab = "Rain fall", main = "Maximum rainfall per month")
+
+lines(KRA.max, type = "o", col = "blue")
+lines(BOL.max, type = "o", col = "green")
+legend(1, 200, legend = c("ADA", "KRA", "BOL"), col= c("red", "blue", "green"), lty=1)
+
+AXM <- c(74.8 , 71.0 , 124.2 , 132.6 , 182.2 , 209.5 , 189 , 162.8 , 110.2 , 180.1 , 112.9 , 87.9 )
+ODA <-  c(59.1 , 79.8 , 127 , 78.3 , 166.9 , 89 , 101.6 , 58.8 , 74.7 , 99.6 , 69.4 , 52.3 )
+BEK <- c(54.3 , 86.8 , 93.6 , 95.7 , 118.4 , 104.4 , 90.2 , 85 , 117.8 , 86 , 53.9 , 154 )
+KDA <- c(43 , 71.2 , 159.6 , 105.8 , 64.7 , 219.9 , 101.8 , 84.8 , 90.6 , 65.5 , 84.7 , 70.4 )
+ABE <- c(65.2 , 72.6 , 91.6 , 97.8 , 79.2 , 145.6 , 77.5 , 87.5 , 121.1 , 97.5 , 51.9 , 67.6 )
+KSI <- c(59.9 , 60.2 , 83.3 ,  96.1 , 125.2 , 145.8 , 90 , 75 , 111.7 , 111.5 , 48.4 , 84 )
[…]

### 2018-05-18 gftests.R

@@ -1,0 +1,13 @@
+ad.test(Uni.oct.n$Data, "pgamma", scale=12.805, shape=0.681)
+
+Bi.oct <- subset(Bi.oct, Data <80)
+
+cvm.test(Semi.jun$Data, "pgamma", scale=30.324, shape=0.529)
+cvm.test(Semi.jun$Data, "pweibull",scale=11.451, shape=0.646)
+
+x <-rgamma(1000, scale = 16.644, shape=0.666)
+
+fit.gamma.semioct <- fitdist(Semi.may$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+Semi.apr.n <- subset(Semi.may, Data <90)
+fit.gamma.semiapr.n <- fitdist(Semi.apr.n$Data, "gamma",  method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+

### 2018-05-18 monthlyPlots.R

@@ -1,0 +1,480 @@
+AXM.jan <- subset(AXM, substr(AXM$Date, 5, 6)=='01')
+AXM.feb <- subset(AXM, substr(AXM$Date, 5, 6)=='02')
+AXM.mar <- subset(AXM, substr(AXM$Date, 5, 6)=='03')
+AXM.apr <- subset(AXM, substr(AXM$Date, 5, 6)=='04')
+AXM.may <- subset(AXM, substr(AXM$Date, 5, 6)=='05')
+AXM.jun <- subset(AXM, substr(AXM$Date, 5, 6)=='06')
+AXM.jul <- subset(AXM, substr(AXM$Date, 5, 6)=='07')
+AXM.aug <- subset(AXM, substr(AXM$Date, 5, 6)=='08')
+AXM.sep <- subset(AXM, substr(AXM$Date, 5, 6)=='09')
+AXM.oct <- subset(AXM, substr(AXM$Date, 5, 6)=='10')
+AXM.nov <- subset(AXM, substr(AXM$Date, 5, 6)=='11')
+AXM.dec <- subset(AXM, substr(AXM$Date, 5, 6)=='12')
+
+AXMmean.jan <- sum(AXM.jan$Data)/30
+AXMmean.feb <- sum(AXM.feb$Data)/30
+AXMmean.mar <- sum(AXM.mar$Data)/30
+AXMmean.apr <- sum(AXM.apr$Data)/30
+AXMmean.may <- sum(AXM.may$Data)/30
+AXMmean.jun <- sum(AXM.jun$Data)/30
+AXMmean.jul <- sum(AXM.jul$Data)/30
+AXMmean.aug <- sum(AXM.aug$Data)/30
+AXMmean.sep <- sum(AXM.sep$Data)/30
+AXMmean.oct <- sum(AXM.oct$Data)/30
+AXMmean.nov <- sum(AXM.nov$Data)/30
+AXMmean.dec <- sum(AXM.dec$Data)/30
+
+AXM.month.rain <- c(AXMmean.jan, AXMmean.feb, AXMmean.mar, AXMmean.apr, AXMmean.may, AXMmean.jun, AXMmean.jul, AXMmean.aug, AXMmean.sep, AXMmean.oct, AXMmean.nov, AXMmean.dec)
+months <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12)
+plot(months, AXM.month.rain, main = "monthly average, AXM")
+
[…]

### 2018-05-18 rainQuantiles.R

@@ -1,0 +1,98 @@
+
+y8392 <- subset(dataNew.norm, substr(dataNew.norm$Date, 1,4)==1983|substr(dataNew.norm$Date, 1,4)==1984|substr(dataNew.norm$Date, 1,4)==1985|substr(dataNew.norm$Date, 1,4)==1986|substr(dataNew.norm$Date, 1,4)==1987|substr(dataNew.norm$Date, 1,4)==1988|substr(dataNew.norm$Date, 1,4)==1989|substr(dataNew.norm$Date, 1,4)==1990|substr(dataNew.norm$Date, 1,4)==1991|substr(dataNew.norm$Date, 1,4)==1992)
+y9302 <- subset(dataNew.norm, substr(dataNew.norm$Date, 1, 4)==1993|substr(dataNew.norm$Date, 1, 4)==1994|substr(dataNew.norm$Date, 1, 4)==1995|substr(dataNew.norm$Date, 1, 4)==1996|substr(dataNew.norm$Date, 1, 4)==1997|substr(dataNew.norm$Date, 1, 4)==1998|substr(dataNew.norm$Date, 1, 4)==1999|substr(dataNew.norm$Date, 1, 4)==2000|substr(dataNew.norm$Date, 1, 4)==2001|substr(dataNew.norm$Date, 1, 4)==2002)
+y0312 <- subset(dataNew.norm, substr(dataNew.norm$Date, 1, 4)==2003|substr(dataNew.norm$Date, 1, 4)==2004|substr(dataNew.norm$Date, 1, 4)==2005|substr(dataNew.norm$Date, 1, 4)==2006|substr(dataNew.norm$Date, 1, 4)==2007|substr(dataNew.norm$Date, 1, 4)==2008|substr(dataNew.norm$Date, 1, 4)==2009|substr(dataNew.norm$Date, 1, 4)==2010|substr(dataNew.norm$Date, 1, 4)==2011|substr(dataNew.norm$Date, 1, 4)==2012)
+
+
+Uni8392 <- subset(y8392, Station=='07008KRA' | Station=='07000BOL' | Station=='07006TLE' | Station=='04003NAV1')
+#6 stations, 21306 pos obs
+Bi8392 <- subset(y8392, Station=='21088ODA' | Station=='22050KDA' | Station=='17009KSI' | Station=='07017HO' | Station=='01032SUN'|Station=='01018WEN')
+#6 stations, 16280 pos obs
[…]

### 2018-05-18 yearlyMaxima.R

@@ -1,0 +1,4 @@
+year <- seq(1983, 2012, 1)
+
+for (i in year) {x <- subset(KRA, substr(data$Date,1, 4)==i)
+                  yearlymax <- max(x$Data)}

### 2018-05-18 DraftReport.tex
future CMIP5 scenarios and density plot on time series

@@
 	\label{R20GCM45}
+\end{figure}
+
+\subsubsection{Different future scenarios}
+Comparing the different carbon scenarios can also give an idea how sensitive the area is to composition changes in the atmosphere.
+The rcp26 seems like a unlikely scenario, wheras rcp45 is a possible outcome and rcp85 is the worst imaginable so far, with hopes that big enough changes will be introduced to avoid it.
+So if we compare the result of these two scenarios, we can see that the two scenarios mostly behave in a similar way, but not everywhere.
+A first interesting observationin fig \ref{R14585} is that the two graphs do not start in the same place, eventhough that is historical data and not future data.
+Secondly, rcp45 is decreasing until around 2060 and flatens out after that, wheras rcp85 has a steady decline for the entire period.
+This is potentially the reson for why that scenario goes down to a lower level, beacuse they are bout around 192 days in 2060.
+Fig \ref{R104585} instead shows a nearly identical behaviour in both graphs.
+Once again, they start at different values (1 day difference) but show an increasing and decreasing bahaviour in the same places except the final years where rcp85 shows a big drop in days which is not at all visible in the rcp45 scenario.
+Fig \ref{R204585} behaves slightly different compared to the other graphs, with rcp45 scenario starting at a larger value then rcp85 and varies more between years.
+But both graphs has the same ranges, indicating that they differ very little from each other.
+So with a large carbon emisson, we seems to have the largest effect on number of rainy days and least on larger rain amounts.
+This should mean that it will be more days with small rain amounts.
+But as seen, the models are not very good at simulating heavy rainfall, which means that there will be a bigger increase in heavy rainfall.
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{R1GCM45mean.pdf}
+		\caption{rcp4.5}
+	\end{subfigure}%
+	\begin{subfigure}{0.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{R1GCM85mean.pdf}
+		\caption{rcp8.5}
+	\end{subfigure}
+	\caption{R1mm for rcp 45 and rcp85 scenario 1980-2100}
+	\label{R14585}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{R10GCM45mean_long.pdf}
+		\caption{rcp4.5}
+	\end{subfigure}%
+	\begin{subfigure}{0.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{R10GCM85mean_long.pdf}
+		\caption{rcp8.5}
+	\end{subfigure}
+	\caption{R10mm for rcp 45 and rcp85 scenario 1980-2100}
+	\label{R104585}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\begin{subfigure}{0.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{R20GCM45mean.pdf}
+		\caption{rcp4.5}
+	\end{subfigure}%
+	\begin{subfigure}{0.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{R20GCM85mean.pdf}
+		\caption{rcp8.5}
+	\end{subfigure}
+	\caption{R20mm for rcp 45 and rcp85 scenario 1980-2100}
+	\label{R204585}
 \end{figure}

### 2018-05-18 Kernel.R

@@ -1,0 +1,7 @@
+df <- data.frame(x=ano, y=annual.rain.20)
+
+g <- ggplot(df, aes(x=ano)) + geom_point(aes(y=annual.rain.20)) +geom_line(aes(y=annual.rain.20)) 
+ggMarginal(g, type = "density", margins  = "y", bw=0.3, fill = "transparent")
+
+plot(density(annual.rain.20, bw=0.1, kernel = "biweight"))
+

### 2018-05-22 MayPresentation.tex
Presentation added

\documentclass{beamer}
\usetheme{Singapore}
\title{Rainfall in Ghana}
\subtitle{A story about extremes}
\author{Jennifer Israelsson}
\institute{University of Reading}
\date{\today}

\begin{document}
	\begin{frame}
	\titlepage
	\end{frame}

\section{Background}

 \begin{frame}
 \frametitle{Ghana}
 \begin{columns}
 	\column{0.5\textwidth}
 	\begin{itemize}
 		\item 4-11$^{\circ}$ N, 3$^{\circ}$W to 1$^{\circ}$E
 		\item Biggest export is agriculture, such as:
 		\begin{itemize}
 			\item Cacao
 			\item Lemons
 			\item Tea
 	\end{itemize}
 	\item Rain depends on the  west African monsoon, controlled by the \textbf{ITCZ} and \textbf{Harmattan}.
 	\item We have 30 years of daily rainfall data from 16 stations
 	\end{itemize}
 	\column{0.5\textwidth}
 	\begin{figure}
 	\includegraphics[width=\linewidth]{ghana-map.jpg}
 	\caption{Map of Africa. https://www.vecteezy.com}
 \end{figure}
 \end{columns}
\end{frame}

\begin{frame}
\frametitle{Inter-tropical convergence zone}
\begin{itemize}
	\item Moves from south of Africa northward to 15$^{\circ}$N in the northern hemisphere summer because of a thermal low pressure.
	\item Cold water is moving northward along the south-west African coast
	\item Pressure gradient between South Atlantic and north Africa which brings cold and moist air, this is the west African monsoon.
	\item Harmattan is when during the northern hemisphere winter, dry and dusty air comes from Sahara.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Rain patterns}
\begin{figure}
	\includegraphics[width=\linewidth]{mapRainmode2.jpg}
\end{figure}
\end{frame}

\section{Distribution fit}

\begin{frame}
\frametitle{Previous used distributions}
\begin{itemize}
	\item Gamma distribution :
	\begin{equation*}
	f(x ; \alpha, \theta) = \frac{x^{\alpha-1} e^{-\frac{x}{\theta}}}{\theta^\alpha \Gamma(\alpha)}, \ x > 0
	\end{equation*}
	\item Lognormal :
	\begin{equation*}
	f(x ; \mu, \sigma)  = \frac{1}{x\sigma \sqrt{2\pi}} e^{- \frac{(ln x - \mu)^2}{2 \sigma ^2}}, \  x > 0
	\end{equation*}
	\item Exponential :
	\begin{equation*}
	f(x ; \lambda) = \lambda e^{-\lambda x}, \  x> 0
	\end{equation*}
	\item Weibull (extreme value) :
	\begin{equation*}
	f(x ; k, \lambda) = \frac{k}{\lambda}\left(\frac{x}{\lambda}\right)^{k-1}e^{(-x/ \lambda)^k}, \ x> 0
	\end{equation*}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{More rain modelling}
\begin{itemize}
	\item Model occurence
	\begin{itemize}
		\item Binomial to model occurence, p mean value of proportion of rainy days in month and location.
		\item Markov chain if we want to model the rainpatterns of a certain length in each month and location.
	\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Fitting distributions}
\begin{itemize}
	\item Used Gamma distribution as first attempt for all months with enough data.
	\begin{itemize}
		\item Excluded December, January, February for north stations.	
	\end{itemize}
	\item \textbf{MLE} method to fit the disribution because it minimizes the variance.
	\item \textbf{QQ plots} to judge how well it fits the distribution.
	\item If a poor fit, tried the other commonly used distributions.
	\item If a good fit for most of the data, tried to cap of the most extreme observations.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Fitting distributions}
\begin{figure}
	\includegraphics[scale=0.12]{UniFitJun.jpg}
	\caption{June for uni mode stations.}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Fitting distributions, capped}
\begin{columns}
	\column{0.5\linewidth}
\begin{figure}
	\includegraphics[width=\linewidth]{UniFitJul.jpg}
	\caption{July for uni mode stations, all data points.}
\end{figure}
\column{0.5\linewidth}
\begin{figure}
	\includegraphics[width=\linewidth]{UniFitJul120.jpg}
	\caption{July for uni mode stations, capped at 120.}
\end{figure}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Fitting distributions, Weibull}
\begin{columns}
	\column{0.5\linewidth}
\begin{figure}
	\includegraphics[scale=0.3]{BiFitAug.pdf}
	\caption{August for bi mode stations, Gamma.}
\end{figure}
\column{0.5\linewidth}
\begin{figure}
	\includegraphics[scale=0.3]{FitBiAugWei.pdf}
	\caption{August for bi mode stations, Weibull.}
\end{figure}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Goodness-of-fit}
\begin{itemize}
	\item Statistical tests to quantify how good or poor fit the chosen distribution is to our data.
	\item GoF test :
	\begin{itemize}
		\item \textbf{Kolmogorov-Smirnov:} 
		\begin{equation*}
		D_n = sup_x |F_n(x) - F(x)|
		\end{equation*}
		\item \textbf{Anderson-Darling:} The test statistics, A, of our ordered data ${x_1, x_2, ..., x_n}$, is defined as,
		\begin{equation*}
		A^2 = -n - S
		\end{equation*}
		where
		\begin{equation*}
		S = \sum_{i=1}^n\frac{2i -1}{n}[\text{ln}(F(x_i)) + \text{ln}(1 - F(x_{n+1-i}) )]
		\end{equation*}
		\item \textbf{Cram\'{e}r-von Mises:} Ordered data ${x_1, x_2,...,x_n}$ ,
		\begin{equation*}
		T = \frac{1}{12n} + \sum_{i=1}^{n}\left[\frac{2i-1}{2n} - F(x_i)\right]^2
		\end{equation*}
	\end{itemize}
\end{itemize}
\end{frame}

\section{Comparing data with CMIP5}

\begin{frame}
\frametitle{Extreme inidicies}
\begin{itemize}
	\item 27 extreme indices set by the \textbf{ETCCDI} (Expert Team on Climate Change Detection and Indices) on temperature and precipitation.
	\item Days with less then 1 mm of rain are classified as dry days.
	\item Mixture of absolute, threshold, duration and percentile indicies. 
	\item Looked at so far:
	\begin{itemize}
		\item \textbf{PRCPTOT :} Total wet day precipitation.
		\item \textbf{R1mm: } Number of wet days.
		\item \textbf{R10mm\textbackslash R20mm:} Number of days with more than 10\textbackslash 20 mm rain.
		\item \textbf{R95p \textbackslash R99p:} Total amount of rain on days with rain above 95 \textbackslash 99 percentile value in reference period.
	\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Absolute indicies, PRCPTOT}
\begin{columns}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{lowessAp.pdf}
		\caption{Average of all the stations}
	\end{figure}
\column{0.5\linewidth}
\begin{figure}
	\includegraphics[width=\linewidth]{PRCPTOTmodes.pdf}
	\caption{Avergage within each group}
\end{figure}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Absolute indicies, PRCPTOT}
\begin{columns}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{PRCPTOTGCMall.pdf}
		\caption{All CMIP5 models}
	\end{figure}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{PRCPTOTGCMmean.pdf}
		\caption{CMIP5 model mean}
	\end{figure}
\end{columns}
Simluates too many days of rain (about 100 days more then our data).
\end{frame}

\begin{frame}
\frametitle{Threshold indicies, R10}
\begin{columns}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{lowessR10.pdf}
		\caption{Average of all the stations}
	\end{figure}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{R10mmodes.pdf}
		\caption{Average within each group}
	\end{figure}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Threshold indicies, R10}
\begin{columns}
\column{0.5\linewidth}
\begin{figure}
	\includegraphics[width=\linewidth]{R10GCMall.pdf}
	\caption{All CMIP5 models}
\end{figure}
\column{0.5\linewidth}
\begin{figure}
	\includegraphics[width=\linewidth]{R10GCMmean.pdf}
	\caption{CMIP5 model mean}
\end{figure}
\end{columns}
Performs resonably well, the mean curve is only slightly over the data mean, but the spread is still massive.
\end{frame}

\begin{frame}
\frametitle{Threshold indicies, R20}
\begin{columns}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{lowessR20.pdf}
		\caption{Average of all the stations}
	\end{figure}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{R20mmodes.pdf}
		\caption{Average within each group}
	\end{figure}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Threshold indicies, R20}
\begin{columns}
\column{0.5\linewidth}
\begin{figure}
	\includegraphics[width=\linewidth]{R20GCMall.pdf}
	\caption{All CMIP5 models}
\end{figure}
\column{0.5\linewidth}
\begin{figure}
	\includegraphics[width=\linewidth]{R20GCMmean.pdf}
	\caption{CMIP5 model mean}
\end{figure}
\end{columns}
Opposite issue, the models simulates too few days instead.
\end{frame}

\begin{frame}
\frametitle{Percentile indicies}
\begin{itemize}
	\item CMIP5 uses 1961-1990 as reference period. 
	\item For my analyses I picked 1984-1993.
	\item To test that the percentile values are robust, calculated 95\% confidence interval using bootstrap method with 5000 samples maximum value included.
	\item 95\% value between (46.2, 48.8).
	\item 99\% value between  (79.3, 88.6).
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Percentile indicies, R95p}
\begin{columns}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{lowessR95.pdf}
		\caption{Mean of all data}
	\end{figure}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{R95GCMmean.pdf}
		\caption{CMIP5 model mean}
	\end{figure}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Percentile indicies, R99p}
\begin{columns}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{lowessR99.pdf}
		\caption{Mean of all data}
	\end{figure}
	\column{0.5\linewidth}
	\begin{figure}
		\includegraphics[width=\linewidth]{R99GCMmean.pdf}
		\caption{CMIP5 model mean}
	\end{figure}
\end{columns}
\end{frame}

\section{Future work}

\begin{frame}
\frametitle{Extreme value statistics}
\begin{itemize}
	\item In extreme value analysis, there can be natural truncation points, but in some cases the data deviates from the Pareto distribution because or data is truncated by other factors e.g rain gauges are too small or a flood barrier is too low. 
	\item Need tools to both investigate if data is truncated and to reconstruct the non-truncated distribution.
	\item These methods can also be used to find the right end point (max value) of the distribution.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Truncations}
There are two types of truncations, rough and light. Define t to be the point from which all values to the right follows a Pareto distribution and T to be the truncation value and t,T $\rightarrow \infty$ as n $\rightarrow \infty$
\begin{itemize}
	\item \textbf{Rough truncation:} $T_n / t_n \rightarrow \beta > 1 $
	\item \textbf{Light truncation:} $T_n / t_n \rightarrow \infty$
\end{itemize}
Tests have been developed to test if T = $\infty$ or finite and to test for light or rough truncation. One can then use TPa (truncated pareto) QQ-plots to graphically investigate if it is a good fit. 
\end{frame}

\begin{frame}
\frametitle{Application}
We will use this theory to:
\begin{itemize}
	\item Investigate if our data is truncated, and in that case try to reconstruct the underlying distribution.
	\item Try and find a right end point on daily rainfall, initially over Ghana, but then apply it to all of Africa to see how it might differ depending on location.
\end{itemize}
\end{frame}

\end{document}

### 2018-05-22 ExtIndicies.R

@@ -1,5 +1,10 @@
 ano = seq(1983,2012)
 
+y8392 <- subset(dataNew.norm, substr(dataNew.norm$Date, 1,4)==1983|substr(dataNew.norm$Date, 1,4)==1984|substr(dataNew.norm$Date, 1,4)==1985|substr(dataNew.norm$Date, 1,4)==1986|substr(dataNew.norm$Date, 1,4)==1987|substr(dataNew.norm$Date, 1,4)==1988|substr(dataNew.norm$Date, 1,4)==1989|substr(dataNew.norm$Date, 1,4)==1990|substr(dataNew.norm$Date, 1,4)==1991|substr(dataNew.norm$Date, 1,4)==1992)
+y9302 <- subset(dataNew.norm, substr(dataNew.norm$Date, 1, 4)==1993|substr(dataNew.norm$Date, 1, 4)==1994|substr(dataNew.norm$Date, 1, 4)==1995|substr(dataNew.norm$Date, 1, 4)==1996|substr(dataNew.norm$Date, 1, 4)==1997|substr(dataNew.norm$Date, 1, 4)==1998|substr(dataNew.norm$Date, 1, 4)==1999|substr(dataNew.norm$Date, 1, 4)==2000|substr(dataNew.norm$Date, 1, 4)==2001|substr(dataNew.norm$Date, 1, 4)==2002)
+y0312 <- subset(dataNew.norm, substr(dataNew.norm$Date, 1, 4)==2003|substr(dataNew.norm$Date, 1, 4)==2004|substr(dataNew.norm$Date, 1, 4)==2005|substr(dataNew.norm$Date, 1, 4)==2006|substr(dataNew.norm$Date, 1, 4)==2007|substr(dataNew.norm$Date, 1, 4)==2008|substr(dataNew.norm$Date, 1, 4)==2009|substr(dataNew.norm$Date, 1, 4)==2010|substr(dataNew.norm$Date, 1, 4)==2011|substr(dataNew.norm$Date, 1, 4)==2012)
+
+
 #annual total precipitation (PRCPTOT)
 annual.prec <- numeric(30)
 for (i in 1:30) {
@@ -22,6 +27,15 @@ for (i in 1:30) {
   annual.rain.10[i] <- length(z$Data)/16
 }
 
+#annual total days >= 20mm (R20mm)
+data.New20 <- subset(dataNew.norm, Data >= 20)
+annual.rain.20 <- numeric(30)
+for (i in 1:30) {
+  a <- subset(data.New20, year==ano[i])
[…]

### 2018-05-22 Kernel.R

@@ -1,7 +1,9 @@
-df <- data.frame(x=ano, y=annual.rain.20)
+df <- data.frame(x=ano, y=annual.rain.10)
 
-g <- ggplot(df, aes(x=ano)) + geom_point(aes(y=annual.rain.20)) +geom_line(aes(y=annual.rain.20)) 
-ggMarginal(g, type = "density", margins  = "y", bw=0.3, fill = "transparent")
+require(ggExtra)
+g <- ggplot(df, aes(x=ano)) + geom_point(aes(y=annual.rain.10)) +geom_line(aes(y=annual.rain.10)) 
+ggMarginal(g, type = "density", margins  = "y", bw=0.5, fill = "transparent")
 
-plot(density(annual.rain.20, bw=0.1, kernel = "biweight"))
+plot(density(annual.rain.10, bw=0.2, kernel = "biweight"))
 
+plot(density(ar20.prop,bw=0.003, kernel="biweight"))

### 2018-05-31 DraftReport.tex
stationary tests added

@@
 84-93 is picked as a reference period instead of 83-92 to avoid the clearly lower values in -83.
+To confirm the robustness of these quantile threshold, I ran a bootstrap with 5000 samples of the same size as the decadal samples.
+After this, I constructed 95\% interval on the threshold values, both with and without forcing the maximum value to be included in the sample.
+Doing this showed that the 95 percentile value is very robust to wether the largest value is included or not, since the confidence interval does not change at all.
+The 99 percentile is slightly higher when forcing the largest value to be included (79.4993, 87.7000 compared to 78.80000, 87.60127 for 03-12 decade).
+The confidence interval is also not as big as one might expect considering it being a very right tail property.
+The 95\% interval is small (47.8,  50.7  for 03-12 decade), hence we can conclude that the value chosen is robust, and does not depend on the specific data that we have.
+
 In figure \ref{R95} both curves seems to exhibit a very similar behaviour, which is a steady increase in the total rainfall on days with heavy rainfall.
@@
 For the quantile based indicies there is instead a positive autocorrelation with a 6 year lag (el ni\~{n}o?).
+
+	\section{Extreme analysis}
+	\subsection{Ground work}
+	In order to use the extreme value theory, in our case \textbf{POT} method, we nee our data to be stationary, to ensure identical, and independent.
+To make data behave independent, we remove the smaller of the two esceedances, if they are closer then a certain number of days from each other.
+This lag is usually set to 3 or 5 days, depending on region and time of year.
+To make sure that the exceedances follows a \textbf{Poisson point process},the standard way is to consider data in cluster of a specific size and then only take the maximum if more then one observation within that cluster exceeds the threshold.
+This makes it important to pick a high enough threshold, since else a lot of our observations will be lost.
+	To ensure stationarity is more diffucult and several approaches are used.
+One is to divide the data into seasons and use different thresholds for each season.
+The issue with this method is that we assume stationarity within each season and therefore can get unrealistic return values if there in fact is a trend in the data.
+If one knows that the data depends on other factors, such as the yield depends on temperature and rain, one can use a Box-Cox transformation on the data and then let the drift and trend parameter depend on these other paramters, so called \textbf{covariates}.
+This however requires that we know of these covariates, can estimate or already have models for how they affect our data and have measurments of these covariates over the same time period as our own data.\\
+
+	{\Large\textbf{HO}}: \textbf{MAM}: Both trend tests gives high p-values, i.e no evidence for a trend in the time series.
+Both unit root tests indicates that the series is stationary/ trend stationary.
+No significant autocorrelation or pattern in lag 1 plot.
+	\textbf{JJA}: Using p=0.1 we reject the null hypothesis and conclude that we have a decresing trend, but not on a 5\% level (MK does not confirm).
+Same result as above for unit root and autocorrelation.
+	\textbf{SON}: No evidence for a trend, unit root or autocorrelation. 
+	\textbf{DJF}: Same as above.
+	None of the stationary seasons rejects the null hypothesis about rough truncation, hence we can use the normal POT method (?).
+The shape parameter for DJF is estimated to be -0.1.
+MAM is estimated to have a shape parameter 0.2 and SON 0.\\
+
+	{\Large\textbf{KRA}}: \textbf{MAM} : Same as above.
+\textbf{JJA}: Decreasing trend on a 10\% level (MK confirms).
+No evidence of a unit root or autocorrelation.
+\textbf{SON}: Strong evidence for an increasing trend.
+Significant autocorrelation for up tp 3 day lag.
+\textbf{DJF}: Evidence on a 10\% level for an increasing trend.No autocorrelation.
+\\
+
+	{\Large\textbf{ADA}}: \textbf{MAM}: Same as above.
+\textbf{JJA}: Evidence on a 5\% level for an incresing trend.
+No evidence for a unit root, but significant autocorrelation up to 11 days lag and some linear pattern in plot of 1 day lag.
+\textbf{SON}: Strong evidence for an increasing trend, but not for a unit root.
+No significant autocorrelation.
+\textbf{DJF}: Evidence on 5\% level for an increase, but no evidence for unit root or autocorrelation.
 

### 2018-05-31 Basic functions.R

@@ -37,27 +37,27 @@ Apr <- subset(data, substr(data$Date,5,6)=='04')
 d.Apr <- as.Date(as.character(Apr$Date), format='%Y%m%d')
 
 #All stations
-ABE <-subset(data.norm, Station=='14067ABE')
-ACC <-subset(data.norm, Station=='23016ACC')
-ADA <-subset(data.norm, Station=='23002ADA')
-AKA <- subset(data.norm, Station=='15076AKA')
-ODA <- subset(data.norm, Station=='21088ODA')
-AXM <- subset(data.norm, Station=='23001AXM')
-BOL <- subset(data.norm, Station=='07000BOL')
-HO <- subset(data.norm, Station=='07017HO')
-KRA <- subset(data.norm, Station=='07008KRA')
-KDA <- subset(data.norm, Station=='22050KDA')
-KSI <- subset(data.norm, Station=='17009KSI')
-NAV1 <- subset(data.norm, Station=='04003NAV1')
-SAL <- subset(data.norm, Station=='23022SAL')
-BEK <- subset(data.norm, Station=='16015BEK')
-SUN <- subset(data.norm, Station=='01032SUN')
-TDI <- subset(data.norm, Station=='23003TDI')
-TEM <- subset(data.norm, Station=='23024TEM')
-TLE <- subset(data.norm, Station=='07006TLE')
-WA <- subset( data.norm, Station == '01013WA')
-WEN <- subset(data.norm, Station=='01018WEN')
-NAV <- subset(data.norm, Station=='04003NAV')
+#ABE <-subset(dataNew, Station=='14067ABE')
+ ACC <-subset(dataNew, Station=='23016ACC')
+ ADA <-subset(dataNew, Station=='23002ADA')
+#AKA <- subset(dataNew, Station=='15076AKA')
+ ODA <- subset(dataNew, Station=='21088ODA')
+ AXM <- subset(dataNew, Station=='23001AXM')
+ BOL <- subset(dataNew, Station=='07000BOL')
+ HO <- subset(dataNew, Station=='07017HO')
+ KRA <- subset(dataNew, Station=='07008KRA')
+ KDA <- subset(dataNew, Station=='22050KDA')
+ KSI <- subset(dataNew, Station=='17009KSI')
+ NAV1 <- subset(dataNew, Station=='04003NAV1')
+ SAL <- subset(dataNew, Station=='23022SAL')
[…]

### 2018-05-31 statsfunction.R

@@ -1,0 +1,81 @@
+summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE, conf.interval=.99) {
+  library(doBy)
+  
+  # New version of length which can handle NA's: if na.rm==T, don't count them
+  length2 <- function (x, na.rm=FALSE) {
+    if (na.rm) sum(!is.na(x))
+    else       length(x)
+  }
+  
+  # Collapse the data
+  formula <- as.formula(paste(measurevar, paste(groupvars, collapse=" + "), sep=" ~ "))
+  datac <- summaryBy(formula, data=data, FUN=c(length2,mean,sd), na.rm=na.rm)
+  
+  # Rename columns
+  names(datac)[ names(datac) == paste(measurevar, ".mean",    sep="") ] <- measurevar
+  names(datac)[ names(datac) == paste(measurevar, ".sd",      sep="") ] <- "sd"
+  names(datac)[ names(datac) == paste(measurevar, ".length2", sep="") ] <- "N"
+  
+  datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
+  
+  # Confidence interval multiplier for standard error
+  # Calculate t-statistic for confidence interval: 
+  # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
+  ciMult <- qt(conf.interval/2 + .5, datac$N-1)
+  datac$ci <- datac$se * ciMult
+  
+  return(datac)
+}
+
+unis <- summarySE(Uni, measurevar = 'Data', groupvars = 'month')
+
+pd <- position_dodge(0.1)
+
+ggplot(unis, aes(x=month, y=Data)) + 
+  geom_errorbar(aes(ymin=Data-ci, ymax=Data+ci), width=.1, position=pd) +
+  geom_line(position=pd) +
+  geom_point(position=pd)
+
+bis <- summarySE(Bi, measurevar = 'Data', groupvars = 'month')
+
+#pd <- position_dodge(0.1)
+
+ggplot(unis, aes(x=month, y=Data)) + 
+  geom_errorbar(aes(ymin=Data-ci, ymax=Data+ci), width=.1) +
+  geom_line() +
+  geom_point()
+
+
+cox.stuart.test =
+  function (x)
+  {
+    method = "Cox-Stuart test for trend analysis"
+    leng = length(x)
+    apross = round(leng) %% 2
+    if (apross == 1) {
+      delete = (length(x)+1)/2
+      x = x[ -delete ] 
+    }
[…]

### 2018-05-31 trPareto.R

@@ -1,0 +1,132 @@
+x_nn <- sort(dataNew.norm$Data, decreasing = TRUE)
+
+x_400n <- x_nn[1:400]
+
+y <- GPDfit(x_nn[1:1000])
+#strong suggestions for gumbel, depending on no of observations used
+#negative shape suggests T finite
+
+r <- gpdSim(model=list(xi=0.0018, mu=0, beta= 26.8876), n=1000)
+r <- rpareto(500, -0.575,140.682)
+
+
+h <-trHill(x_100n, logk = TRUE, plot = TRUE)
+trEndpoint(x_100n, gamma = h$gamma, plot=TRUE)
+
+#must use Hill to estimate gamma because need gamma to be positive
+GH <- genHill(x_nn[1:700], gamma=H$gamma, plot=TRUE, col="blue", ylim=c(-0.2,0.4))
+
+
+H <- Hill(x_700n, plot=TRUE, col="blue")
+# Add EPD estimator
+EPD(x_nn, add=TRUE, col="orange", lty=2)
+legend("bottomright", c("Hill","EPD"), col=c("blue","orange"), lty=1:2)
+
+#Quantile estimation
+p <- 0.0005
+x <- x_nn[5:400]
+
+#must use Hill to estimate gamma because need gamma to be positive
+H <- Hill(x)
+GH <- genHill(x, gamma=H$gamma)
+QuantGH(x, gamma=GH$gamma,p=p, plot= TRUE, col="blue")
+g <- GPDmle(x)
+QuantGPD(x, gamma=g$gamma, sigma = g$sigma, p=p, add=TRUE, col="green")
+M <- Moment(x)
+QuantMOM(x, gamma=M$gamma, p=p, add=TRUE, col="orange")
+
+
+#test for truncation
+tM <- trMLE(x,start = c(1,1), plot=FALSE)
+tTM <- trTestMLE(x, gamma=tM$gamma, tau=tM$tau, alpha = 0.05, plot = TRUE, main = "Test for truncation")
+
+#trPareto estimation
+x <- x_nn[1:600]
+
+tH <- trHill(x, r = 1, tol = 1e-08, maxiter = 100, logk = FALSE,plot = FALSE, add = FALSE, main = "Estimates of the EVI")
+tD <- trDT(x, r = 1, gamma=tH$gamma, plot = FALSE, add = FALSE, main = "Estimates of DT")
[…]

### 2018-06-01 DraftReport.tex
code and files added for time series of ADA,KRA,HO

@@
 	\subsection{Ground work}
-	In order to use the extreme value theory, in our case \textbf{POT} method, we nee our data to be stationary, to ensure identical, and independent.
-To make data behave independent, we remove the smaller of the two esceedances, if they are closer then a certain number of days from each other.
+	In order to use the extreme value theory, in our case \textbf{POT} method, we need our data to be stationary, to ensure identical, and independent.
+To make data behave independent, we remove the smaller exceedances, if they are more then one closer then a certain number of days from each other.
 This lag is usually set to 3 or 5 days, depending on region and time of year.
@@
 This makes it important to pick a high enough threshold, since else a lot of our observations will be lost.
-	To ensure stationarity is more diffucult and several approaches are used.
+	To ensure stationarity is more difficult and several approaches are used.
 One is to divide the data into seasons and use different thresholds for each season.
@@
 This however requires that we know of these covariates, can estimate or already have models for how they affect our data and have measurments of these covariates over the same time period as our own data.\\
+	To investigate stationarity in our time serie, we first applied a decomposition on both the full time series and then on each season.
+This to look at trends both on an annual time scale but also on a seasonal one.
+By decomposing even the seasons we can remove influence on seasonality within each season, i.e we usually have a lot more rain in September then in November.
+To statisically check for stationarity, we used the \textbf{Augmented dickey-Fuller test} (ADF).
+Many uses \textbf{ManKendell}, but since this test only uses non ties and we have loads of zeroes in several seasons, it will not be very reliable (?).
+
+	\subsubsection{Time series analysis}
+	{\Large\textbf{HO}}: Figure \ref{tsHO} shows the trends for years and seasons.
+The only season that is showing some sort of trend is DJF, which seems to have a slight increasing trend.
+However, the scale is so small,that it is more or less neglectibe.
+SON is not showing a clear trend, but an increased volatility in the trend after 1995.
+There is a big volatilty in the JJA trend, suggesting that the rain in that season depends on some other process.
+MAM is the most steady season.
+All time series rejects the null hypothesis for the ADF test, hence suggesting that there is no unit root, i.e stationary.
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.6\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsHO.pdf}
+			\caption{Full time series}
+		\end{subfigure}
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=\linewidth]{tsHOmam.pdf}
+			\caption{MAM}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsHOjja.pdf}
+			\caption{JJA}
+		\end{subfigure}
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=\linewidth]{tsHOson.pdf}
+			\caption{SON}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsHOdjf.pdf}
+			\caption{DJF}
+		\end{subfigure}%
+		\caption{Decomposition of time series for HO}
+		\label{tsHO}
+	\end{figure}
+
+	{\Large\textbf{ADA}}: In figure \ref{tsADA} we can once again see that there is no clear trend on the annual time series, which the ADF test confirms.
+MAM has again the smoothest trend series with only one big dip around 1988, and there seems to be a decrease in the most extreme values.
+JJA is similarly to HO showing periodic behaviour indicating that there is some underlying process effecting the rain in this season.
+Cannot really detect a trend.
+SON shows a fairly flat trend graph, but an increase in extreme observations after some dry years between 1990-2000.
+DJF is also showing an increse in the amplitude of extreme values even if the number of extreme values is not clearly increasing.
+The tend graph is not flat mostly due to the fact that we nearly only have zero values.
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.6\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsADA.pdf}
+			\caption{Full time series}
+		\end{subfigure}
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=\linewidth]{tsADAmam.pdf}
+			\caption{MAM}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsADAjja.pdf}
+			\caption{JJA}
+		\end{subfigure}
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=\linewidth]{tsADAson.pdf}
+			\caption{SON}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsADAdjf.pdf}
+			\caption{DJF}
+		\end{subfigure}%
+		\caption{Decomposition of time series for ADA}
+		\label{tsADA}
+	\end{figure}
+
+	{\Large\textbf{KRA}}: Comparing fig \ref{tsKRA} with the two previous, we can see a lot of similarities.
+There is once again no trend in the full time series and it is somewhere between the behaviour of HO and ADA, with HO being much smoother and ADA more volitile.
+But for this station MAM and JJA are similarly smooth.
+MAM seems to have an increase in amplitude in extreme values, which is not visible in JJA.
+SON instead looks like it is having less exteme values, wheras DJF clearly has an increase both in number and amplitude in extreme values.
+All time series rejects the null hypothesis for the ADF test.
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.6\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsKRA.pdf}
+			\caption{Full time series}
+		\end{subfigure}
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=\linewidth]{tsKRAmam.pdf}
+			\caption{MAM}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsKRAjja.pdf}
+			\caption{JJA}
+		\end{subfigure}
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=\linewidth]{tsKRAson.pdf}
+			\caption{SON}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\centering
+			\includegraphics[width=1\linewidth]{tsKRAdjf.pdf}
+			\caption{DJF}
+		\end{subfigure}%
+		\caption{Decomposition of time series for KRA}
+		\label{tsKRA}
+	\end{figure}
 

### 2018-06-01 WorkingReport.tex
code and files added for time series of ADA,KRA,HO

@@
 This flow brings cold and moist oceanic air which is the reason for rain, so the area with maximum rainfall is where we have thickest air masses, i.e closest to the coast.
-North of the ITCSZ conditions are genrally cloudless and dry and south of the zone we find the most cloud cover and maximum rain.
-Dry years are coused by: retardation of the northward movement of the south-west monsoon, a southward displacemnt of the intertropical discontinuity,the near-equatorial trough and the xone of maximum surface pressure and anoalously cold water to the nort-west of the line linking SW West Africa nd NE Brasil.
+North of the ITCSZ conditions are generally cloudless and dry and south of the zone we find the most cloud cover and maximum rain.
+Dry years are caused by: retardation of the northward movement of the south-west monsoon, a southward displacement of the intertropical discontinuity,the near-equatorial trough and the zone of maximum surface pressure and anomalously cold water to the nort-west of the line linking SW West Africa and NE Brasil.
 	During the northern hemisphere winter, West Africa is instead under the influence of  north-easterly trade winds which bring dry and stable air masses, often containing dust particles.
@@
 These two zones can be as far aprat as 1000 km.
-	\textbf{Squall lines} is a linear system of many thunderstorms or clouds that behave as one and therefore can live for 13-15 h instead of just a couple of hourse.
-They can be hundreds of km long and 30 km wide and are characterised by their explosive growth, rapid propagation and their convex leading ege(?).
-They peak in the afternoon and usually form west of mountains which implies that surface heating and orographic  effects creates them.
 	 In west Africa, the African easterly jet seems to have a big influence on the initiation of them.
@@
 This can be done in numerous of ways, linear regression being one or running the model and changing one variable at a time.
+		\item \textbf{Squall lines} is a linear system of many thunderstorms or clouds that behave as one and therefore can live for 13-15 h instead of just a couple of hourse.
+They can be hundreds of km long and 30 km wide and are characterised by their explosive growth, rapid propagation and their convex leading ege(?).
+They peak in the afternoon and usually form west of mountains which implies that surface heating and orographic  effects creates them
 	\end{itemize}

### 2018-06-01 timeSeries.R

@@ -1,0 +1,25 @@
+#Pick one station and split up the seasons
+x <- KRA
+
+DJF <- subset(x, month==1| month==2|month==12)
+MAM <- subset(x, month==3|month==4 |month==5)
+JJA <- subset(x, month==6|month==7|month==8)
+SON <- subset(x, month==9|month==10|month==11)
+
+#transform into timeseries
+x.ts <- ts(x$Data, frequency = 365.25, start=c(1983,1,1))
+DJF.ts <- ts(DJF$Data,frequency = 90.25, start=1983)
+MAM.ts <- ts(MAM$Data,frequency = 91, start=1983)
+JJA.ts <- ts(JJA$Data,frequency = 92, start=1983)
+SON.ts <- ts(SON$Data,frequency = 91, start=1983)
+
+#decompose and test for stationarity
+decompX <- stl(x.ts, s.window = "periodic")
+plot(decompX)
+adf.test(x.ts)
+
+y <- JJA.ts
+
+decomp <- stl(y, s.window = 91)
+plot(decomp)
+adf.test(y)

### 2018-06-01 trPareto.R

@@ -66,24 +66,6 @@ JJA <- subset(x, month==6|month==7|month==8)
 SON <- subset(x, month==9|month==10|month==11)
 
 #plot time-serie/trend
-x.ts <- ts(x$Data, frequency = 365.25, start=c(1983,1,1))
-plot(x.ts, main=unique(x$Station), ylab="rain amount (mm)")
-
-DJF.ts <- ts(DJF$Data,frequency = 90.25, start=1983)
-plot(DJF.ts, main="DJF", ylab="rain(mm)")
-lines(lowess(DJF.ts, f=0.0001), col = 6)
-
-MAM.ts <- ts(MAM$Data,frequency = 91, start=1983)
-plot(MAM.ts, main="MAM", ylab="rain(mm)")
-lines(lowess(MAM.ts, f=0.0001), col = 6)
-
-JJA.ts <- ts(JJA$Data,frequency = 92, start=1983)
-plot(MAM.ts, main="JJA", ylab="rain(mm)")
-lines(lowess(JJA.ts, f=0.0001), col = 6)
-
-SON.ts <- ts(SON$Data,frequency = 91, start=1983)
-plot(SON.ts, main="SON", ylab="rain(mm)")
-lines(lowess(SON.ts, f=0.0001), col = 6)
 
 #test for trend and stationarity
 y <- MAM

### 2018-06-07 DraftReport.tex
extreme fitting with and without truncation updated

@@
 Nearly all months in Bi is having a negative kurtosis wheras for the other only half are negative.
-So many of the sets are more lightly tailed then a normal distribution.\\
+So many of the sets are more lightly tailed than a normal distribution.\\
 
@@
 By decomposing even the seasons we can remove influence on seasonality within each season, i.e we usually have a lot more rain in September then in November.
-To statisically check for stationarity, we used the \textbf{Augmented dickey-Fuller test} (ADF).
-Many uses \textbf{ManKendell}, but since this test only uses non ties and we have loads of zeroes in several seasons, it will not be very reliable (?).
+This since looking at only one month at a time would lead to too little data to work with for the extreme analysis.
+To statisically check for stationarity, we aggregated the data into months for the full time series and yearly for the seasons.
+We then apply the Mann-Kendall test on the seasons, the seasonal Mann-Kendall on the full time series and lastly Sen's slope test on both the full and seasonal to confirm the MK test.
 
@@
 The only season that is showing some sort of trend is DJF, which seems to have a slight increasing trend.
-However, the scale is so small,that it is more or less neglectibe.
+However, the scale is so small,that it is more or less neglectible.
 SON is not showing a clear trend, but an increased volatility in the trend after 1995.
@@
 All time series rejects the null hypothesis for the ADF test, hence suggesting that there is no unit root, i.e stationary.
+	The Mann-Kendall test confirms that DJF has an increasing trend on a 5\% level, but none of the other seasons and the sesonal MK test does not reject the null hypothesis.
 	\begin{figure}[H]
@@
 DJF is also showing an increse in the amplitude of extreme values even if the number of extreme values is not clearly increasing.
-The tend graph is not flat mostly due to the fact that we nearly only have zero values.
+The trend graph is not flat mostly due to the fact that we nearly only have zero values.
+	Here the Mann-Kendall test gives the exact same result as for the other stations, there is a significant increase in DJF but not in any of the other seasons or for the overall time series.
+
 	\begin{figure}[H]
@@
 All time series rejects the null hypothesis for the ADF test.
+
+	Aggregating the data from daily to monthly, we can use the Mann-Kendall and seasonal Mann-Kendall on the data to quantitativly look at trends.
+The only season that shows a significant trend is DJF, which one should expect considering the increase in heavy rain.
+Both the seasonal and all other MK tests are insignificant.
 	\begin{figure}[H]
@@
 \textbf{DJF}: Evidence on 5\% level for an increase, but no evidence for unit root or autocorrelation.
+
+	\subsection{POT method}
+	Concluding that no seasons except DJF shows an increasing trend, we can apply the POT straight onto our data.
+One can of course argue wheater all data in a season is i.i.d since there still is a seasonality trend in there, but for the moment we choose to negect this.
+Fitting an extreme distribution to DJF is of no point, since we barely have any data to start with, which means that all of the positive data points are to be considered extreme cases.
+But since we see an increase both in number and intensity, their behaviour is interesting to model, perhaps in a linear regression setting.
+
+	For the fitting of GPD (Pareto, $\xi > 0$), the following method will be used:
+	\begin{itemize}
+		\item Pick out the top 30\% of the data above 1 mm.
+		\item Smooth it to make it continuose.
+		\item Test for rough truncation. Choose suitable formulas depending on result.
+		\item Find the shape parameter using Hill, MLE and moments.
+		\item Fit the data to a GPD and confirm the suitability of the fit with a QQ-plot.
+		\item Find the end point and high quantiles. 
+	\end{itemize}
+
+	\subsubsection{ADA}
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.5\textwidth]{monAveADA.png}
+		\caption{Monthly averages}
+	\end{figure}
+
+	For ADA, the maximum rainfall usually falls in June, so therefore the maximum value shoudl be found in that season.
+For now DJF will not be analysed since even if there is a trend in it, extremes in that season are still not real extremes.
+Analysing MAM, we cannot reject the null hypothesis about light truncation, so we will use the traditional POT method.
+Looking at ME plot, it looks like k=100 can be a suitable threshold which is roughly 20\% of the positive data.
+All methods agree that the shape parameter is around 0, MLE and MOM becomes negative after k=80.
+This suggests that there exists a finite end point.
+Applying the POT method with 100 exceedances gives a nearly perfect fit with $\xi = -0.11$, which gives a much better fit then the MLE fitted values, potentially because of the declustering procedure.
+Using p=0.0001 gives a quantile value of around 180.
+Plotting the exceedances show no trend pattern.
+	\begin{figure}
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{shapeADAMAM.pdf}
+			\caption{Shape estimation, MAM}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{potADAMAM.pdf}
+			\caption{POT fitting, $\xi = -0.11$}
+		\end{subfigure}
+	\end{figure}
+
+	For JJA, the shape paratmer seems to be positive, hence we can use all the tests for a pareto distribution.
+We do not reject the null for the MLE truncation test but for the other one.
+k=120 seems to be a suitable value judging on the ME plot, which is roughly 25\% of the positive data.
+Hill estimates $\xi = 0.5$ and allt the other $\xi = 0.25$.
+The regular POT method estimates $\xi = 0.06$ but also provides a relatively poor fit.
+No trend in the exceedances.
+Using the truncated end point estimation, we get a theoretical end point around 270, wheras if we use normal quantile estimation, p=0.0001 gives a value or 500, which clearly demonstrates the advantage of using the truncated Pareto distribution instead.
+
+	\begin{figure}[H]
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{shapeADAJJA.pdf}
+			\caption{Shape estimation, JJA}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{MEadaJJA.pdf}
+			\caption{ME plot}
+		\end{subfigure}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.3\textwidth}
+			\includegraphics[width=\linewidth]{potADAJJA.pdf}
+			\caption{POT estimation}
+		\end{subfigure}%
+		\begin{subfigure}{0.3\textwidth}
+			\includegraphics[width=\linewidth]{QQadaJJAPOTshape.pdf}
+			\caption{QQ plot,POT shape estimator}
+		\end{subfigure}%
+		\begin{subfigure}{0.3\textwidth}
+			\includegraphics[width=\linewidth]{TPaQQadaJJA.pdf}
+			\caption{TPa }
+		\end{subfigure}%
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{endTadaJJA.pdf}
+			\caption{End point estimaton}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{extQuanADAJJA.pdf}
+			\caption{p=0.0001}
+		\end{subfigure}
+	\end{figure}
+
+	SON behaves rather unpredictable.
+The ME plot has a negative slope which would indicate a positive shape parameter, but all estimators except Hill estimates it to be around 0.
+For the MLE test, we reject the null hypothesis after k=50 but not before and we reject the null for the other test if k$<$50.
+The POT plot is once again a near perfect fit, and the residual plots looks straight, whereas the general QQ is not straight at all.
+The truncated pareto also seems to give a good fit.
+The end point is estimated to be 120.
+If using p=0.0001, there is massive volatility in the moment graph, wheras the rest are stable around 170.
+They agree with the endpoint for k upto 30.
+
+	\begin{figure}[H]
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{shapeADASON.pdf}
+			\caption{Shape estimation, SON}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{MEadaSON.pdf}
+			\caption{ME plot, SON}
+		\end{subfigure}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{trMLEadaSON.pdf}
+			\caption{MLE estimation}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{trTestADASON.pdf}
+			\caption{Hill estimator}
+		\end{subfigure}
+	\caption{Truncation test}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{potADASON.pdf}
+			\caption{POT estimation}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{TPaQQadaSON.pdf}
+			\caption{TPa }
+		\end{subfigure}%
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{endTadaSON.pdf}
+			\caption{End point estimaton}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{extQuanADASON.pdf}
+			\caption{p=0.0001}
+		\end{subfigure}
+	\end{figure}
 

### 2018-06-07 timeSeries.R

@@ -1,5 +1,5 @@
 #Pick one station and split up the seasons
-x <- KRA
+x <- ADA
 
 DJF <- subset(x, month==1| month==2|month==12)
 MAM <- subset(x, month==3|month==4 |month==5)
@@ -23,3 +23,62 @@ y <- JJA.ts
 decomp <- stl(y, s.window = 91)
 plot(decomp)
 adf.test(y)
+
+#Mann-Kendall on yearly data
+ano = seq(1983,2012)
+
+agg.year <- numeric(30)
+for (i in 1:30) {
+  k <- subset(DJF, year==ano[i])
+  agg.year[i] <- sum(k$Data)
+ }
+
+MannKendall(agg.year)
+
+#seasonal MK
+agg.mon <- numeric(360)
+for (i in 1:30) {
+  k <- subset(x, year==ano[i])
+  
+  k.jan <- subset(k, substr(k$Date, 5, 6)=='01')
+  k.feb <- subset(k, substr(k$Date, 5, 6)=='02')
+  k.mar <- subset(k, substr(k$Date, 5, 6)=='03')
+  k.apr <- subset(k, substr(k$Date, 5, 6)=='04')
+  k.may <- subset(k, substr(k$Date, 5, 6)=='05')
+  k.jun <- subset(k, substr(k$Date, 5, 6)=='06')
+  k.jul <- subset(k, substr(k$Date, 5, 6)=='07')
+  k.aug <- subset(k, substr(k$Date, 5, 6)=='08')
+  k.sep <- subset(k, substr(k$Date, 5, 6)=='09')
+  k.oct <- subset(k, substr(k$Date, 5, 6)=='10')
+  k.nov <- subset(k, substr(k$Date, 5, 6)=='11')
+  k.dec <- subset(k, substr(k$Date, 5, 6)=='12')
+  
+  agg.mon[12*(i-1)+1] <- sum(k.jan$Data)
+  agg.mon[12*(i-1) +2] <- sum(k.feb$Data)
+  agg.mon[12*(i-1)+3] <- sum(k.mar$Data)
+  agg.mon[12*(i-1) +4] <- sum(k.apr$Data)
+  agg.mon[12*(i-1)+5] <- sum(k.may$Data)
+  agg.mon[12*(i-1)+6] <- sum(k.jun$Data)
+  agg.mon[12*(i-1)+7] <- sum(k.jul$Data)
[…]

### 2018-06-07 trPareto.R

@@ -1,3 +1,4 @@
+require(ReIns)
 x_nn <- sort(dataNew.norm$Data, decreasing = TRUE)
 
 x_400n <- x_nn[1:400]
@@ -58,7 +59,7 @@ trQuantMLE(x, gamma=tM$gamma, tau=tM$tau, DT=tD$DT, p=p, Y = TRUE, add = TRUE, c
 
 #Truncation analysis for stations
 #pick station for analysis
-x <- KRA
+x <- ADA
 
 DJF <- subset(x, month==1| month==2|month==12)
 MAM <- subset(x, month==3|month==4 |month==5)
@@ -68,7 +69,7 @@ SON <- subset(x, month==9|month==10|month==11)
 #plot time-serie/trend
 
 #test for trend and stationarity
-y <- MAM
+y <- SON
   
 trend.test(y$Data)
 cox.stuart.test(y$Data)
@@ -85,30 +86,73 @@ x_t <- y$Data[2:n]
 x_t1 <- y$Data[1:n-1]
 plot(x_t1, x_t)
 
-#POT method,declustering = 5 days
+#Pick out the tail data
+require(evir)
 y.pos <- subset(y, Data >= 1)
-n <- floor(length(y.pos$Data)*0.1)
-pot(y$Data, nextremes = n, run = 5)
+n <- floor(length(y.pos$Data)*0.3)
+MeanExcess(y.pos$Data, plot=TRUE, k=TRUE)
+n <- 80
 
 y_nn <- sort(y$Data, decreasing = TRUE)
 z <- y_nn[1:n]
 
 #check for duplicated values, smoothing (make it cts). m=2 => +0.025, m=3 => 1/6, 1/2, 5/6
 z[duplicated(z) | duplicated(z, fromLast=TRUE)]
-which(z==46.6)
 
 #if duplicates at position z[a]
-a <- 34
-z[a-1] <- z[a-1] + 0.025
-z[a] <- z[a] - 0.025
+a <-which(z==18.2)[1]
+# 2 ties
+z[a+1] <- z[a+1] - 0.025
+z[a] <- z[a] + 0.025
 
-tM <- trMLE(z,start = c(1,1), plot=FALSE)
+#3 ties
+z[a] <- z[a] + 0.16
+z[a+2] <- z[a+2] - 0.16
+
+#Test for rough truncation
+tM <- trMLE(z,start = c(0.1,1))
+tT <- trTest(z, plot=TRUE) #must have >0 shape parameter
[…]

### 2018-06-14 weathergenerator.R
rain generator with 1 step markov chain, both with and without threshold selection. Needs trPareto to do threshold selestion

@@ -1,0 +1,150 @@
+x <- ADA
+
+DJF <- subset(x, month==1| month==2|month==12)
+MAM <- subset(x, month==3|month==4 |month==5)
+JJA <- subset(x, month==6|month==7|month==8)
+SON <- subset(x, month==9|month==10|month==11)
+
+e <- MAM$Data
+
+#1 step Markov chain
+num_rn = 0
+for (i in 1:(length(e)-1)) {
+  if ((e[i+1]>=1) & (e[i]<1)){
+    num_rn <- num_rn + 1
+  }
+}
+p_rn <- num_rn/length(e)
+
+num_nr = 0
+for (i in 1:(length(e)-1)) {
+  if ((e[i+1]<1) & (e[i]>=1)){
+    num_nr <- num_nr + 1
+  }
+}
+p_nr <- num_nr/length(e)
+
+num_nn = 0
+for (i in 1:(length(e)-1)) {
+  if ((e[i+1]<1) & (e[i]<1)){
+    num_nn <- num_nn + 1
+  }
+}
+p_nn <- num_nn/length(e)
+
+num_rr = 0
+for (i in 1:(length(e)-1)) {
+  if ((e[i+1]>=1) & (e[i]>=1)){
+    num_rr <- num_rr + 1
+  }
+}
+p_rr <- num_rr/length(e)
+
+#probability of rain
+prr <- num_rr/(num_nr + num_rr)
+prn <- num_rn/(num_rn + num_nn)
+
+rainydays <- subset(e, e>=1)
+
+require(fitdistrplus)
+fit <- fitdist(rainydays, distr = "gamma", method = "mle", lower= c(0,0), start = list(scale = 1, shape = 1))
+qqcomp(fit)
+summary(fit)
+plot(fit)
+
+k<- length(e)
+occurence <- numeric(k)
+
+rainonfirstday <- 1
+occurence[1] <- rainonfirstday
+
+number <- runif(k, 0, 1)
+for (i in 2:(k-1)) {
+  if((occurence[i-1]>0) & (number[i] < prr))
+  {occurence[i] <- 1}
+  else if ((occurence[i-1]==0) & (number[i] < prn))
+    {occurence[i] <- 1}
+  else {occurence[i] <- 0}
+}
+
+amount <- rgamma(k, shape = fit$estimate[2], scale=fit$estimate[1])
+
+for (i in 1:length(amount)){
[…]

### 2018-06-14 DraftReport.tex
simulation section on ADA MAM added

@@
 
-	For ADA, the maximum rainfall usually falls in June, so therefore the maximum value shoudl be found in that season.
+	For ADA, the maximum rainfall usually falls in June, so therefore the maximum value should be found in that season.
 For now DJF will not be analysed since even if there is a trend in it, extremes in that season are still not real extremes.
 Analysing MAM, we cannot reject the null hypothesis about light truncation, so we will use the traditional POT method.
-Looking at ME plot, it looks like k=100 can be a suitable threshold which is roughly 20\% of the positive data.
+Looking at ME plot, it looks like k=50 can be a suitable threshold which is roughly 10\% of the positive data.
 All methods agree that the shape parameter is around 0, MLE and MOM becomes negative after k=80.
 This suggests that there exists a finite end point.
-Applying the POT method with 100 exceedances gives a nearly perfect fit with $\xi = -0.11$, which gives a much better fit then the MLE fitted values, potentially because of the declustering procedure.
-Using p=0.0001 gives a quantile value of around 180.
+Applying the POT method with 50 exceedances gives a nearly perfect fit with $\xi = -0.08$.
 Plotting the exceedances show no trend pattern.
 	\begin{figure}
-		\begin{subfigure}{0.5\textwidth}
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shEstADAMAM.pdf}
+			\caption{Shape estimation}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{MEadaMAM.pdf}
+			\caption{ME plot}
+		\end{subfigure}
+		\begin{subfigure}{0.33\textwidth}
 			\includegraphics[width=\linewidth]{shapeADAMAM.pdf}
-			\caption{Shape estimation, MAM}
-		\end{subfigure}%
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{potADAMAM.pdf}
-			\caption{POT fitting, $\xi = -0.11$}
-		\end{subfigure}
-	\end{figure}
-
-	For JJA, the shape paratmer seems to be positive, hence we can use all the tests for a pareto distribution.
-We do not reject the null for the MLE truncation test but for the other one.
-k=120 seems to be a suitable value judging on the ME plot, which is roughly 25\% of the positive data.
-Hill estimates $\xi = 0.5$ and allt the other $\xi = 0.25$.
-The regular POT method estimates $\xi = 0.06$ but also provides a relatively poor fit.
-No trend in the exceedances.
-Using the truncated end point estimation, we get a theoretical end point around 270, wheras if we use normal quantile estimation, p=0.0001 gives a value or 500, which clearly demonstrates the advantage of using the truncated Pareto distribution instead.
-
-	\begin{figure}[H]
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{shapeADAJJA.pdf}
-			\caption{Shape estimation, JJA}
-		\end{subfigure}%
-		\begin{subfigure}{0.5\textwidth}
+			\caption{Shape estimation}
+		\end{subfigure}
+	\caption{MAM}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot50ADAMAM.pdf}
+			\caption{POT estimation $\xi = -0.08$}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{extQuanADAMAM.pdf}
+			\caption{Quantile estimation}
+		\end{subfigure}
+	\caption{MAM}
+	\end{figure}
+
+	For JJA, we once again use the largest 50 data points, and we appear to have a positive shape parameter, with the Hill estimator once again giving a much larger value compared to the other.
+The slightly lower value of MLE could be since the data has not been declustered.
+Using \textbf{POT} method we once again get a positive shape value, similar to the other estimator, wheras if we use the \textbf{GPDfit} function, also using MLE, we get a negative shape value, both if we decluster the data or not.
+I cannot find any reason for this, but since all other estimators get it to be positive, that seems most resonable.
+This would imply that we do not have a finite right end point.
+
+	\begin{figure}[H]
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shEstADAJJA.pdf}
+			\caption{Shape estimation}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
 			\includegraphics[width=\linewidth]{MEadaJJA.pdf}
@@
 		\end{subfigure}
-	\end{figure}
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.3\textwidth}
-			\includegraphics[width=\linewidth]{potADAJJA.pdf}
-			\caption{POT estimation}
-		\end{subfigure}%
-		\begin{subfigure}{0.3\textwidth}
-			\includegraphics[width=\linewidth]{QQadaJJAPOTshape.pdf}
-			\caption{QQ plot,POT shape estimator}
-		\end{subfigure}%
-		\begin{subfigure}{0.3\textwidth}
-			\includegraphics[width=\linewidth]{TPaQQadaJJA.pdf}
-			\caption{TPa }
-		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shapeADAJJA.pdf}
+			\caption{Shape estimation}
+		\end{subfigure}
+		\caption{JJA}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot50ADAJJA.pdf}
+			\caption{POT estimation, $\xi = 0.153$}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{QQGFadaJJA.pdf}
+			\caption{QQ plot,MLE estimation}
+		\end{subfigure}
+		\caption{JJA}
 	\end{figure}
@@
 			\includegraphics[width=\linewidth]{extQuanADAJJA.pdf}
-			\caption{p=0.0001}
-		\end{subfigure}
+			\caption{Quantile estimation}
+		\end{subfigure}
+	\caption{JJA}
 	\end{figure}
@@
 	\begin{figure}[H]
-		\begin{subfigure}{0.5\textwidth}
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shEstADASON.pdf}
+			\caption{Shape estimation}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{MEadaSON.pdf}
+			\caption{ME plot}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
 			\includegraphics[width=\linewidth]{shapeADASON.pdf}
+			\caption{Shape estimation}
+		\end{subfigure}
+		\caption{SON}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot40ADASON.pdf}
+			\caption{40 points,$\xi = -0.06$}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot45ADASON.pdf}
+			\caption{45 points,$\xi = -0.204$}
+		\end{subfigure}
+		\caption{POT estimation, SON}
+	\end{figure}
+
+		\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{extQuan40ADASON.pdf}
+			\caption{40 points}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{extQuan45ADASON.pdf}
+			\caption{45 points}
+		\end{subfigure}
+		\caption{Extreme quantiles, SON}
+	\end{figure}
+
+	\subsubsection{HO}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.5\textwidth]{monAveHO.png}
+		\caption{Monthly averages}
+	\end{figure}
+
+	Depending if we take the number of extremes to be 50 or 100 in MAM, we get very different behaviours, so therefore both cases will be investigated.
+Nothing interesting is happening around the 50th biggest value.
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shEstHOMAM.pdf}
+			\caption{Shape estimation, MAM}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{MEhoMAM.pdf}
+			\caption{ME plot, MAM}
+		\end{subfigure}
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shapeHOMAM.pdf}
+			\caption{Shape estimates, MAM}
+		\end{subfigure}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{trMLEhoMAM.pdf}
+			\caption{MLE estimation}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{trTestHOMAM.pdf}
+			\caption{Hill estimator}
+		\end{subfigure}
+		\caption{Truncation test}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot90HOMAM.pdf}
+			\caption{POT $\xi = 0.006$, 90 extremes}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{extQuanHOMAM.pdf}
+			\caption{p=$10^{-5}$}
+		\end{subfigure}
+	\end{figure}
+
+	Since the truncation end point only can be estimated for values where we do not reject light truncation, we cannot use it.
+The top 55 points fits well to the POT model and there is a stable value for the shape parameter.
+This would indicate that we have a theoretical maximum value of 300, but if we instead take k $>60$ it is instead 240.
+
+	\begin{figure}[H]
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shEstHOJJA.pdf}
+			\caption{Shape estimation, JJA}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{MEhoJJA.pdf}
+			\caption{ME plot, JJA}
+		\end{subfigure}%
+			\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shapeHOJJA.pdf}
+			\caption{Shape estimation, JJA}
+		\end{subfigure}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot40HOJJA.pdf}
+			\caption{POT $\xi = 0.04$, 40 extremes}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot60HOJJA.pdf}
+			\caption{POT $\xi = -0.06$, 60 extremes}
+		\end{subfigure}
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot100HOJJA.pdf}
+			\caption{POT, $\xi = 0.05$, 100 extremes }
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{pot110HOJJA.pdf}
+			\caption{POT, $\xi = -0.06$, 110 extremes }
+		\end{subfigure}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{exccHOJJA.pdf}
+			\caption{100 largest values}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{extQuanHOJJA.pdf}
+			\caption{p=$10^{-5}$}
+		\end{subfigure}
+	\end{figure}
+
+	JJA has a very similar pattern to MAM, which rejects light truncation for k $>$ 40 but not before.
+The endpoint estimator can once again not be used, since it only gives values for very small k.
+But here there seems to be an increasing trend in the extremes, which means that our maximum estimate will not be accurate for future scenarios.
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shEstHOSON.pdf}
 			\caption{Shape estimation, SON}
 		\end{subfigure}%
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{MEadaSON.pdf}
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{MEhoSON.pdf}
 			\caption{ME plot, SON}
 		\end{subfigure}
-	\end{figure}
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{trMLEadaSON.pdf}
-			\caption{MLE estimation}
-		\end{subfigure}%
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{trTestADASON.pdf}
-			\caption{Hill estimator}
-		\end{subfigure}
-	\caption{Truncation test}
-	\end{figure}
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{potADASON.pdf}
-			\caption{POT estimation}
-		\end{subfigure}%
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{TPaQQadaSON.pdf}
-			\caption{TPa }
-		\end{subfigure}%
-	\end{figure}
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{endTadaSON.pdf}
-			\caption{End point estimaton}
-		\end{subfigure}%
-		\begin{subfigure}{0.5\textwidth}
-			\includegraphics[width=\linewidth]{extQuanADASON.pdf}
-			\caption{p=0.0001}
-		\end{subfigure}
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{shapeHOSON.pdf}
+			\caption{Shape estimates, SON}
+		\end{subfigure}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{pot40HOSON.pdf}
+			\caption{POT $\xi = 0.19$, 40 extremes}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{pot90HOSON.pdf}
+			\caption{POT $\xi = 0.009$, 90 extremes}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{extQuanHOSON.pdf}
+			\caption{p=$10^{-5}$}
+		\end{subfigure}
+	\end{figure}
+
+	\section{Simulated rain analysis}
+	\subsection{ADA}
+	\subsubsection{MAM}
+	Simulating from only a gamma distribution often gives a very similar plot to the observed data, some simulaions misses values above 100. The mean is within 0.15 of the correct one.
+
+	\begin{table}[H]
+		\centering
+	\begin{tabular}{|c|c|c|c|c|}
+		\hline
+		Simulation & k & Thres & Trunc & POT shape \\
+		\hline
+		1 & 50 & 44.2 &  NO & 0.04 \\
+		2 & 50 & 40 & YES & -0.34 \\
+		3 & 75 & 35.2 & YES & -0.19 \\
+		4 & 50 & 41.7 & NO & 0.09 \\
+	\end{tabular}
+	\caption{Only gamma, no truncation}
+	\end{table}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{wgADAMAMme.pdf}
+			\caption{ME plot}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{wgADAMAMshape.pdf}
+			\caption{Shape estimation}
+		\end{subfigure}
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{wgADAMAMpot.pdf}
+			\caption{POT, $\xi =-0.57$}
+		\end{subfigure}%
+		\begin{subfigure}{0.5\textwidth}
+			\includegraphics[width=\linewidth]{wgADAMAMq}
+			\caption{Quantiles}
+		\end{subfigure}
+	\caption{Only gamma, no truncation}
+	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{wgADAMAMttNT.pdf}
+			\caption{No truncation}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{pot90HOSON.pdf}
+			\caption{Truncated at 90\%}
+		\end{subfigure}%
+		\begin{subfigure}{0.33\textwidth}
+			\includegraphics[width=\linewidth]{extQuanHOSON.pdf}
+			\caption{Truncated at 99\%}
+		\end{subfigure}
+	\caption{Truncation}
 	\end{figure}
